---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# stgeomx

<!-- badges: start -->
[![CRAN status](https://www.r-pkg.org/badges/version/stgeomx)](https://CRAN.R-project.org/package=stgeomx)
[![R-CMD-check](https://github.com/mkiyer/stgeomx/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/mkiyer/stgeomx/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The stgeomx package should be used to assist in the processing and analysis of Nanostring Geomx RNA data. It has a novel background correction function (qq) and a quantile normalization method that correct for background noise in data generated by GeoMx. This package can be used to load GeoMx data from xslx or tsv files, merge datasets together, visualize the quality of the data then set thresholds to filter out probes or ROIs, and apply the package's background correction and normalization to the data. See our manuscript for more information about the package and see our vignette for step-by-step instructions on how to use the package.

## Installation

You can install and load the development version of stgeomx from CRAN with:

``` {r installation CRAN}
install.packages("path/to/your/package", repos = NULL, type = "source")
library("stgeomx")
```

You can install and load the development version of stgeomx from GitHub with:
``` {r installation github}
# install.packages("devtools")
# devtools::install_github("mkiyer/stgeomx")
```
