---
title: "soa"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{soa}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Setup

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(tidyverse)
library(readxl)
library(ggrepel)
library(pheatmap)

# bioconductor packages
library(factoextra)

```

## Vignette-specific functions

```{r vignette functions, include=FALSE}

add_study_field <- function(ds, study) {
  s <- ds$samples
  s$study <- study
  s <- s %>% mutate(patient = sprintf("%s_%s", study, patient),
                    slide = sprintf("%s_%s", study, slide),
                    roi = sprintf("%s_%s", study, roi),
                    aoi = sprintf("%s_%s", study, aoi))
  colnames(ds$counts) <- s$aoi
  return(list(samples = s, meta = ds$meta, counts = ds$counts))
}

save_pheatmap_pdf <- function(x, filename, width=7, height=7) {
  stopifnot(!missing(x))
  stopifnot(!missing(filename))
  pdf(filename, width=width, height=height)
  grid::grid.newpage()
  grid::grid.draw(x$gtable)
  dev.off()
}

```

## Input files and parameter settings

```{r input files, include=FALSE}

# input directories
working_dir <- "/Users/madisoncaldwell/Desktop/stgeomx"

# EDITED (Madison)
data_dir <- "/Users/madisoncaldwell/Desktop/research/data"

# geomx data file location
geomx_dir <- file.path(working_dir, "geomx")

#
# spatial organ atlas datasets
#

soa_panc_geomx_xlsx <- file.path(data_dir, "spatial_organ_atlas_pancreas.xlsx")
soa_panc_geomx_annot_xlsx <- file.path(data_dir, "pancreas_annot.xlsx")
soa_kidney_geomx_xlsx <- file.path(data_dir, "spatial_organ_atlas_kidney.xlsx")
soa_kidney_geomx_annot_xlsx <- file.path(data_dir, "kidney_annot.xlsx")
soa_brain_geomx_xlsx <- file.path(data_dir, "spatial_organ_atlas_brain.xlsx")
soa_brain_geomx_annot_xlsx <- file.path(data_dir, "brain_annot.xlsx")
soa_colon_geomx_xlsx <- file.path(data_dir, "spatial_organ_atlas_colon.xlsx")
soa_colon_annot_xlsx <- file.path(data_dir, "colon_annot.xlsx")
soa_ln_geomx_xlsx <- file.path(data_dir, "spatial_organ_atlas_lymph_node.xlsx")
soa_ln_geomx_annot_xlsx <- file.path(data_dir, "lymph_node_annot.xlsx")

#
# output directory settings
#
results_dir <- file.path(working_dir, "results")
plot_dir <- file.path(results_dir, "plots")
# create output directories
for (d in c(results_dir, plot_dir)) {
  if (!dir.exists(d)) {
    dir.create(d)
  }
}

#
# filtering parameter settings
#
# gene probes must be greater than the Nth quantile of the noise/background
# probes to be considered 'detectable'
geomx_negprobe_lod_quantile = 0.9
# minimum number of total counts for each area-of-interest (AOI)
geomx_min_counts = 50000
# signal-to-noise area under the curve (snAUC) threshold for each AOI
geomx_min_auc = 0.60
# AOI must have at least this fraction of all genes detectable
geomx_aoi_min_frac_expr = 0.2
# a gene must be 'detectable' in at least this fraction of the AOIs to be kept
geomx_gene_min_frac_expr = 0.15

#
# background correction / normalization parameter settings
#
# background (noise) correction methods:
# bgsub: subtract a constant noise value from every gene
# qq: models and subtracts background noise relative to gene probe count
# none: no background correction
bgcorrect_method = "qq"
# kernel density estimation is used to model a smooth noise distribution
# from the small set of background noise probes. modifying the bandwidth 
# parameter will affect how "smooth" the background distribution model is
bgcorrect_bw_adjust = 2
# ensure that gene probes below this quantile in the background noise 
# distribution are zero. A value of 0.5 means that the gene is 
# indistinguishable from background (50% likelihood of being noise)
bgcorrect_bg_quant = 0.5
#
# normalization method options:
# qn: quantile normalization (recommended)
# rle: deseq2 normalization
# cpm: counts per million
# q3: upper quartile normalization
# tmm: edgeR TMM normalization
# none: no normalization
norm_method = "q3"

```


# stgeomx

## Read, process, and merge SOA geomx data

```{r soa panc geomx data, include=FALSE}

# read data
ds_panc <- stgeomx::read_xlsx(soa_panc_geomx_xlsx)
ds_ln <- stgeomx::read_xlsx(soa_ln_geomx_xlsx)
ds_kidney <- stgeomx::read_xlsx(soa_kidney_geomx_xlsx)
ds_colon <- stgeomx::read_xlsx(soa_colon_geomx_xlsx)
ds_brain <- stgeomx::read_xlsx(soa_brain_geomx_xlsx)
ds <- ds_panc

# read and join sample annotation
annot_panc <- read_excel(soa_panc_geomx_annot_xlsx)
annot_ln <- read_excel(soa_ln_geomx_annot_xlsx)
annot_kidney <- read_excel(soa_kidney_geomx_annot_xlsx)
annot_colon <- read_excel(soa_colon_annot_xlsx)
annot_brain <- read_excel(soa_brain_geomx_annot_xlsx)
annot <- annot_panc

# combine samples and annotations
ds$samples <- ds$samples %>%
  inner_join(annot, by="SegmentDisplayName")
ds_panc$samples <- ds_panc$samples %>%
  inner_join(annot_panc, by="SegmentDisplayName")
ds_ln$samples <- ds_ln$samples %>%
  inner_join(annot_ln, by="SegmentDisplayName")
ds_kidney$samples <- ds_kidney$samples %>%
  inner_join(annot_kidney, by="SegmentDisplayName")
ds_colon$samples <- ds_colon$samples %>%
  inner_join(annot_colon, by="SegmentDisplayName")
ds_brain$samples <- ds_brain$samples %>%
  inner_join(annot_brain, by="SegmentDisplayName")

# add study fields to aoi
ds <- add_study_field(ds, "pancreas")
ds_panc <- add_study_field(ds_panc, "pancreas")
ds_ln <- add_study_field(ds_ln, "ln")
ds_kidney <- add_study_field(ds_kidney, "kidney")
ds_colon <- add_study_field(ds_colon, "colon")
ds_brain <- add_study_field(ds_brain, "brain")

# merge datasets (all annotated data)
ds <- stgeomx::merge(ds, ds_ln)
ds <- stgeomx::merge(ds, ds_colon)
ds <- stgeomx::merge(ds, ds_brain)
ds <- stgeomx::merge(ds, ds_kidney)

# process dataset
ds_panc <- stgeomx::preprocess(ds_panc, geomx_negprobe_lod_quantile)
ds_ln <- stgeomx::preprocess(ds_ln, geomx_negprobe_lod_quantile)
ds_colon <- stgeomx::preprocess(ds_colon, geomx_negprobe_lod_quantile)
ds_kidney <- stgeomx::preprocess(ds_kidney, geomx_negprobe_lod_quantile)
ds_brain <- stgeomx::preprocess(ds_brain, geomx_negprobe_lod_quantile)
ds <- stgeomx::preprocess(ds, geomx_negprobe_lod_quantile)

# merge probes when there is >1 probe per gene
ds_panc <- stgeomx::merge_probes(ds_panc)
ds_ln <- stgeomx::merge_probes(ds_ln)
ds_colon <- stgeomx::merge_probes(ds_colon)
ds_kidney <- stgeomx::merge_probes(ds_kidney)
ds_brain <- stgeomx::merge_probes(ds_brain)
ds <- stgeomx::merge_probes(ds)

# set filtering thresholds
ds_panc <- stgeomx::set_thresholds(ds_panc,
                              min_counts = geomx_min_counts,
                              min_auc = geomx_min_auc,
                              aoi_min_frac_expr = geomx_aoi_min_frac_expr,
                              gene_min_frac_expr = geomx_gene_min_frac_expr)
ds_ln <- stgeomx::set_thresholds(ds_ln,
                              min_counts = geomx_min_counts,
                              min_auc = geomx_min_auc,
                              aoi_min_frac_expr = geomx_aoi_min_frac_expr,
                              gene_min_frac_expr = geomx_gene_min_frac_expr)
ds_colon <- stgeomx::set_thresholds(ds_colon,
                              min_counts = geomx_min_counts,
                              min_auc = geomx_min_auc,
                              aoi_min_frac_expr = geomx_aoi_min_frac_expr,
                              gene_min_frac_expr = geomx_gene_min_frac_expr)
ds_kidney <- stgeomx::set_thresholds(ds_kidney,
                              min_counts = geomx_min_counts,
                              min_auc = geomx_min_auc,
                              aoi_min_frac_expr = geomx_aoi_min_frac_expr,
                              gene_min_frac_expr = geomx_gene_min_frac_expr)
ds_brain <- stgeomx::set_thresholds(ds_brain,
                              min_counts = geomx_min_counts,
                              min_auc = geomx_min_auc,
                              aoi_min_frac_expr = geomx_aoi_min_frac_expr,
                              gene_min_frac_expr = geomx_gene_min_frac_expr)
ds <- stgeomx::set_thresholds(ds,
                              min_counts = geomx_min_counts,
                              min_auc = geomx_min_auc,
                              aoi_min_frac_expr = geomx_aoi_min_frac_expr,
                              gene_min_frac_expr = geomx_gene_min_frac_expr)

```

## QC Figures: Nucleic count and surface area vs background noise
```{r nucleic count and surface area v bg}
bg_correlations_dir <- file.path(plot_dir, "bg_correlations")
p <- ggplot(ds$samples, aes(x=.data$AOISurfaceArea, y=.data$bg_cpm, color=.data$study)) +
  geom_point() +
  geom_smooth(method = "lm", col = "red") +
  scale_x_log10() +
  xlab("Area of Sample") +
  ylab("Background Noise")
p

w <- 9
h <- 4
f <- file.path(bg_correlations_dir, "surface_area_bg")
ggsave(paste0(f, ".png"), p, width=w, height=h)
ggsave(paste0(f, ".pdf"), p, width=w, height=h)

p <- ggplot(ds$samples, aes(x=.data$AOINucleiCount, y=.data$bg_cpm, color=.data$study)) +
  geom_point() +
  geom_smooth(method = "lm", col = "red") +
  xlab("Area of Sample") +
  ylab("Background Noise")
p

w <- 9
h <- 4
f <- file.path(bg_correlations_dir, "nuclei_count_bg")
ggsave(paste0(f, ".png"), p, width=w, height=h)
ggsave(paste0(f, ".pdf"), p, width=w, height=h)
```

## QC Figures: AOI filtering, results stratified by different parameters
```{r qc filter aoi, stratification}
ds_func <- ds_brain

width <- 12
height <- 8
 
# bg_auc
p <- ggplot(ds_func$samples, aes(x=patient, y=bg_auc, color=subset)) +
  geom_boxplot(outlier.shape = NA, alpha=0.5) +
  geom_point(aes(group=path_aoi, fill=path_aoi), pch=21, color="black", size=2, alpha=0.7,
             position = position_jitterdodge(jitter.width=0.05, dodge.width=0.5)) +
  geom_hline(yintercept=0.5) +
  geom_hline(yintercept=geomx_min_auc, linetype="dashed", color="red") +
  scale_color_manual(values = pals::cols25()) +
  scale_fill_manual(values = pals::cols25()) +
  facet_wrap(~study, scales="free_x", nrow=1) +
  theme_bw() +
  theme(axis.text.x = element_text(color = "black", angle = 90, vjust = 0.5, hjust=1)) +
  theme(legend.position = "bottom")
p
f <- file.path(plot_dir, "/updated_qc/qc_study_patient_bg_auc")
ggsave(paste0(f, ".pdf"), plot=p, width=width, height=height)
ggsave(paste0(f, ".png"), plot=p, width=width, height=height)
 
# num_counts
p <- ggplot(ds_func$samples, aes(x=patient, y=num_counts+1, color=subset)) +
  geom_boxplot(outlier.shape = NA, alpha=0.5) +
  geom_point(aes(group=path_aoi, fill=path_aoi), pch=21, color="black", size=2, alpha=0.7,
             position = position_jitterdodge(jitter.width=0.05, dodge.width=0.5)) +
  geom_hline(yintercept=geomx_min_counts, linetype="dashed", color="red") +
  scale_y_log10() +
  scale_color_manual(values = pals::cols25()) +
  scale_fill_manual(values = pals::cols25()) +
  facet_wrap(~study, scales="free_x", nrow=1) +
  theme_bw() +
  theme(axis.text.x = element_text(color = "black", angle = 90, vjust = 0.5, hjust=1)) +
  theme(legend.position = "bottom")
p
f <- file.path(plot_dir, "/updated_qc/qc_study_patient_counts")
ggsave(paste0(f, ".pdf"), plot=p, width=width, height=height)
ggsave(paste0(f, ".png"), plot=p, width=width, height=height)
 
# counts across histopathology
p <- ggplot(ds_func$samples, aes(x=subset, y=num_counts+1, fill=path)) +
  geom_boxplot(outlier.shape = NA, alpha=0.5) +
  geom_point(pch=21, color="black", size=2, alpha=0.7,
             position = position_jitterdodge(jitter.width=0.05, dodge.width=0.5)) +
  geom_hline(yintercept=geomx_min_counts, linetype="dashed", color="red") +
  scale_y_log10() +
  scale_color_manual(values = pals::cols25()) +
  facet_wrap(~study, scales="free_x", nrow=1) +
  theme_bw() +
  theme(axis.text.x = element_text(color = "black", angle = 90, vjust = 0.5, hjust=1)) +
  theme(legend.position = "none") 
p
f <- file.path(plot_dir, "/updated_qc/qc_path_counts")
ggsave(paste0(f, ".pdf"), plot=p, width=10, height=5)
ggsave(paste0(f, ".png"), plot=p, width=10, height=5)
 
# bg_auc across histopathology
p <- ggplot(ds_func$samples, aes(x=subset, y=bg_auc, fill=path)) +
  geom_boxplot(outlier.shape = NA, alpha=0.5) +
  geom_point(pch=21, color="black", size=2, alpha=0.7,
             position = position_jitterdodge(jitter.width=0.05, dodge.width=0.5)) +
  geom_hline(yintercept=geomx_min_auc, linetype="dashed", color="red") +
  scale_color_manual(values = pals::cols25()) +
  facet_wrap(~study, scales="free_x", nrow=1) +
  theme_bw() +
  theme(axis.text.x = element_text(color = "black", angle = 90, vjust = 0.5, hjust=1)) +
  theme(legend.position = "none") 
p
f <- file.path(plot_dir, "/updated_qc/qc_path_bg_auc")
ggsave(paste0(f, ".pdf"), plot=p, width=10, height=5)
ggsave(paste0(f, ".png"), plot=p, width=10, height=5)
 
# counts vs auc: subset
p <- ggplot(ds_func$samples, aes(x=num_counts+1, y=bg_auc)) +
  geom_point(aes(color=subset)) +
  geom_vline(xintercept=geomx_min_counts, linetype="dashed", color="red") +
  geom_hline(yintercept=geomx_min_auc, linetype="dashed", color="red") +
  scale_x_log10() +
  scale_color_manual(values = pals::cols25()) +
  theme_bw() +
  facet_wrap(~study)
p
f <- file.path(plot_dir, "/updated_qc/qc_scatter_num_counts_vs_auc_study_subset.pdf")
ggsave(paste0(f, ".pdf"), plot=p, width=8, height=6)
ggsave(paste0(f, ".png"), plot=p, width=8, height=6)

# counts vs auc: region
p <- ggplot(ds_func$samples, aes(x=num_counts+1, y=bg_auc)) +
  geom_point(aes(color=region)) +
  geom_vline(xintercept=geomx_min_counts, linetype="dashed", color="red") +
  geom_hline(yintercept=geomx_min_auc, linetype="dashed", color="red") +
  scale_x_log10() +
  scale_color_manual(values = pals::cols25()) +
  theme_bw() +
  facet_wrap(~study)
p
f <- file.path(plot_dir, "/updated_qc/qc_scatter_num_counts_vs_auc_study_region.pdf")
ggsave(paste0(f, ".pdf"), plot=p, width=8, height=6)
ggsave(paste0(f, ".png"), plot=p, width=8, height=6)

# counts vs auc: slide (sample)
p <- ggplot(ds_func$samples, aes(x=num_counts+1, y=bg_auc)) +
  geom_point(aes(color=slide)) +
  geom_vline(xintercept=geomx_min_counts, linetype="dashed", color="red") +
  geom_hline(yintercept=geomx_min_auc, linetype="dashed", color="red") +
  scale_x_log10() +
  scale_color_manual(values = pals::cols25()) +
  theme_bw() +
  facet_wrap(~study)
p
f <- file.path(plot_dir, "/updated_qc/qc_scatter_num_counts_vs_auc_study_sample.pdf")
ggsave(paste0(f, ".pdf"), plot=p, width=8, height=6)
ggsave(paste0(f, ".png"), plot=p, width=8, height=6)

```


## QC Figures: AOI filtering plots

```{r qc filter aoi}
aoi_filtering_dir <- file.path(plot_dir, "aoi_filtering")

p_merged <- stgeomx::plot_aoi_filter(ds, 
                              min_counts=geomx_min_counts,
                              min_auc=geomx_min_auc,
                              min_frac_expr=geomx_aoi_min_frac_expr)
p_merged
w <- 9
h <- 4
f <- file.path(aoi_filtering_dir, "qc_aoi_filtering_merged")
ggsave(paste0(f, ".png"), p_merged, width=w, height=h)
ggsave(paste0(f, ".pdf"), p_merged, width=w, height=h)

p_panc <- stgeomx::plot_aoi_filter(ds_panc, 
                              min_counts=geomx_min_counts,
                              min_auc=geomx_min_auc,
                              min_frac_expr=geomx_aoi_min_frac_expr)
p_panc
w <- 9
h <- 4
f <- file.path(aoi_filtering_dir, "qc_aoi_filtering_panc")
ggsave(paste0(f, ".png"), p_panc, width=w, height=h)
ggsave(paste0(f, ".pdf"), p_panc, width=w, height=h)

p_ln <- stgeomx::plot_aoi_filter(ds_ln, 
                              min_counts=geomx_min_counts,
                              min_auc=geomx_min_auc,
                              min_frac_expr=geomx_aoi_min_frac_expr)
p_ln
w <- 9
h <- 4
f <- file.path(aoi_filtering_dir, "qc_aoi_filtering_ln")
ggsave(paste0(f, ".png"), p_ln, width=w, height=h)
ggsave(paste0(f, ".pdf"), p_ln, width=w, height=h)

p_kidney <- stgeomx::plot_aoi_filter(ds_kidney, 
                              min_counts=geomx_min_counts,
                              min_auc=geomx_min_auc,
                              min_frac_expr=geomx_aoi_min_frac_expr)
p_kidney
w <- 9
h <- 4
f <- file.path(aoi_filtering_dir, "qc_aoi_filtering_kidney")
ggsave(paste0(f, ".png"), p_kidney, width=w, height=h)
ggsave(paste0(f, ".pdf"), p_kidney, width=w, height=h)

p_colon <- stgeomx::plot_aoi_filter(ds_colon, 
                              min_counts=geomx_min_counts,
                              min_auc=geomx_min_auc,
                              min_frac_expr=geomx_aoi_min_frac_expr)
p_colon
w <- 9
h <- 4
f <- file.path(aoi_filtering_dir, "qc_aoi_filtering_colon")
ggsave(paste0(f, ".png"), p_colon, width=w, height=h)
ggsave(paste0(f, ".pdf"), p_colon, width=w, height=h)

p_brain <- stgeomx::plot_aoi_filter(ds_brain, 
                              min_counts=geomx_min_counts,
                              min_auc=geomx_min_auc,
                              min_frac_expr=geomx_aoi_min_frac_expr)
p_brain
w <- 9
h <- 4
f <- file.path(aoi_filtering_dir, "qc_aoi_filtering_brain")
ggsave(paste0(f, ".png"), p_brain, width=w, height=h)
ggsave(paste0(f, ".pdf"), p_brain, width=w, height=h)

```

## QC Figures: Gene Filtering plots

```{r gene filtering}
gene_filtering_dir <- file.path(plot_dir, "gene_filtering")

p_merged <- stgeomx::plot_gene_filter(ds, geomx_gene_min_frac_expr)
p_merged
w <- 7
h <- 7
f <- file.path(gene_filtering_dir, "qc_gene_filtering_merged")
ggsave(paste0(f, ".png"), p_merged, width=w, height=h)
ggsave(paste0(f, ".pdf"), p_merged, width=w, height=h)

p_panc <- stgeomx::plot_gene_filter(ds_panc, geomx_gene_min_frac_expr)
p_panc
w <- 7
h <- 7
f <- file.path(gene_filtering_dir, "qc_gene_filtering_panc")
ggsave(paste0(f, ".png"), p_panc, width=w, height=h)
ggsave(paste0(f, ".pdf"), p_panc, width=w, height=h)

p_ln <- stgeomx::plot_gene_filter(ds_ln, geomx_gene_min_frac_expr)
p_ln
w <- 7
h <- 7
f <- file.path(gene_filtering_dir, "qc_gene_filtering_ln")
ggsave(paste0(f, ".png"), p_ln, width=w, height=h)
ggsave(paste0(f, ".pdf"), p_ln, width=w, height=h)

p_colon <- stgeomx::plot_gene_filter(ds_colon, geomx_gene_min_frac_expr)
p_colon
w <- 7
h <- 7
f <- file.path(gene_filtering_dir, "qc_gene_filtering_colon")
ggsave(paste0(f, ".png"), p_colon, width=w, height=h)
ggsave(paste0(f, ".pdf"), p_colon, width=w, height=h)

p_brain <- stgeomx::plot_gene_filter(ds_brain, geomx_gene_min_frac_expr)
p_brain
w <- 7
h <- 7
f <- file.path(gene_filtering_dir, "qc_gene_filtering_brain")
ggsave(paste0(f, ".png"), p_brain, width=w, height=h)
ggsave(paste0(f, ".pdf"), p_brain, width=w, height=h)

p_kidney <- stgeomx::plot_gene_filter(ds_kidney, geomx_gene_min_frac_expr)
p_kidney
w <- 7
h <- 7
f <- file.path(gene_filtering_dir, "qc_gene_filtering_kidney")
ggsave(paste0(f, ".png"), p_kidney, width=w, height=h)
ggsave(paste0(f, ".pdf"), p_kidney, width=w, height=h)

```

## Apply filters, then BG correction and normalization

```{r background correction}

correction_normalization_dir <- file.path(plot_dir, "bgcorrection_normalization")

# MERGED
# background correction must use original unfiltered matrix
fds <- stgeomx::bgcorrect(ds, method=bgcorrect_method, 
                          bw.adjust=bgcorrect_bw_adjust, 
                          bg.quant=bgcorrect_bg_quant)
# normalize
fds <- stgeomx::normalize(fds, norm_method)

# plot expression distribution
p <- stgeomx::plot_expr_dist(fds)
p
w <- 8
h <- 8
f <- file.path(correction_normalization_dir, "qc_expr_dist_merged")
ggsave(paste0(f, ".png"), p, width=w, height=h)
ggsave(paste0(f, ".pdf"), p, width=w, height=h)

# plot background correction (before and after)
p <- stgeomx::plot_bgcorrect(fds)
p
w <- 8
h <- 8
f <- file.path(correction_normalization_dir, "qc_bgcorrect_merged")
ggsave(paste0(f, ".png"), p, width=w, height=h)
ggsave(paste0(f, ".pdf"), p, width=w, height=h)

# background correction must use original unfiltered matrix
fds <- stgeomx::bgcorrect(ds, method=bgcorrect_method,
                          bw.adjust=bgcorrect_bw_adjust,
                          bg.quant=bgcorrect_bg_quant)
# apply filters before normalizing
fds <- stgeomx::apply_filters(fds)
# normalize
fds <- stgeomx::normalize(fds, norm_method)

# filtering stats
nrow(ds$samples)
nrow(fds$samples)
table(ds$meta$keep)

# final normalized gene expression matrix
log2_norm_cpm_merged <- log2(fds$x + 1)

# PANCREAS
# background correction must use original unfiltered matrix
fds_panc <- stgeomx::bgcorrect(ds_panc, method=bgcorrect_method, 
                          bw.adjust=bgcorrect_bw_adjust, 
                          bg.quant=bgcorrect_bg_quant)
# normalize
fds_panc <- stgeomx::normalize(fds_panc, norm_method)

# plot expression distribution
p <- stgeomx::plot_expr_dist(fds_panc)
p
w <- 8
h <- 8
f <- file.path(correction_normalization_dir, "qc_expr_dist_panc")
ggsave(paste0(f, ".png"), p, width=w, height=h)
ggsave(paste0(f, ".pdf"), p, width=w, height=h)

# plot background correction (before and after)
p <- stgeomx::plot_bgcorrect(fds_panc)
p
w <- 8
h <- 8
f <- file.path(correction_normalization_dir, "qc_bgcorrect_panc")
ggsave(paste0(f, ".png"), p, width=w, height=h)
ggsave(paste0(f, ".pdf"), p, width=w, height=h)

# background correction must use original unfiltered matrix
fds_panc <- stgeomx::bgcorrect(ds_panc, method=bgcorrect_method,
                          bw.adjust=bgcorrect_bw_adjust,
                          bg.quant=bgcorrect_bg_quant)
# apply filters before normalizing
fds_panc <- stgeomx::apply_filters(fds_panc)
# normalize
fds_panc <- stgeomx::normalize(fds_panc)

# filtering stats
nrow(ds$samples)
nrow(fds_panc$samples)
table(ds$meta$keep)

# final normalized gene expression matrix
log2_norm_cpm_panc <- log2(fds_panc$x + 1)

# LYMPH NODE
# background correction must use original unfiltered matrix
fds_ln <- stgeomx::bgcorrect(ds_ln, method=bgcorrect_method, 
                          bw.adjust=bgcorrect_bw_adjust, 
                          bg.quant=bgcorrect_bg_quant)
# normalize
fds_ln <- stgeomx::normalize(fds_ln, norm_method)

# plot expression distribution
p <- stgeomx::plot_expr_dist(fds_ln)
p
w <- 8
h <- 8
f <- file.path(correction_normalization_dir, "qc_expr_dist_ln")
ggsave(paste0(f, ".png"), p, width=w, height=h)
ggsave(paste0(f, ".pdf"), p, width=w, height=h)

# plot background correction (before and after)
p <- stgeomx::plot_bgcorrect(fds_ln)
p
w <- 8
h <- 8
f <- file.path(correction_normalization_dir, "qc_bgcorrect_ln")
ggsave(paste0(f, ".png"), p, width=w, height=h)
ggsave(paste0(f, ".pdf"), p, width=w, height=h)

# background correction must use original unfiltered matrix
fds_ln <- stgeomx::bgcorrect(ds_ln, method=bgcorrect_method,
                          bw.adjust=bgcorrect_bw_adjust,
                          bg.quant=bgcorrect_bg_quant)
# apply filters before normalizing
fds_ln <- stgeomx::apply_filters(fds_ln)
# normalize
fds_ln <- stgeomx::normalize(fds_ln, norm_method)

# filtering stats
nrow(ds$samples)
nrow(fds_ln$samples)
table(ds$meta$keep)

# final normalized gene expression matrix
log2_norm_cpm_ln <- log2(fds_ln$x + 1)

# KIDNEY
# background correction must use original unfiltered matrix
fds_kidney <- stgeomx::bgcorrect(ds_kidney, method=bgcorrect_method, 
                          bw.adjust=bgcorrect_bw_adjust, 
                          bg.quant=bgcorrect_bg_quant)
# normalize
fds_kidney <- stgeomx::normalize(fds_kidney, norm_method)

# plot expression distribution
p <- stgeomx::plot_expr_dist(fds_kidney)
p
w <- 8
h <- 8
f <- file.path(correction_normalization_dir, "qc_expr_dist_kidney")
ggsave(paste0(f, ".png"), p, width=w, height=h)
ggsave(paste0(f, ".pdf"), p, width=w, height=h)

# plot background correction (before and after)
p <- stgeomx::plot_bgcorrect(fds_kidney)
p
w <- 8
h <- 8
f <- file.path(correction_normalization_dir, "qc_bgcorrect_kidney")
ggsave(paste0(f, ".png"), p, width=w, height=h)
ggsave(paste0(f, ".pdf"), p, width=w, height=h)

# background correction must use original unfiltered matrix
fds_kidney <- stgeomx::bgcorrect(ds_kidney, method=bgcorrect_method,
                          bw.adjust=bgcorrect_bw_adjust,
                          bg.quant=bgcorrect_bg_quant)
# apply filters before normalizing
fds_kidney <- stgeomx::apply_filters(fds_kidney)
# normalize
fds_kidney <- stgeomx::normalize(fds_kidney, norm_method)

# filtering stats
nrow(ds$samples)
nrow(fds_kidney$samples)
table(ds$meta$keep)

# final normalized gene expression matrix
log2_norm_cpm_kidney <- log2(fds_kidney$x + 1)

# COLON
# background correction must use original unfiltered matrix
fds_colon <- stgeomx::bgcorrect(ds_colon, method=bgcorrect_method, 
                          bw.adjust=bgcorrect_bw_adjust, 
                          bg.quant=bgcorrect_bg_quant)
# normalize
fds_colon <- stgeomx::normalize(fds_colon, norm_method)

# plot expression distribution
p <- stgeomx::plot_expr_dist(fds_colon)
p
w <- 8
h <- 8
f <- file.path(correction_normalization_dir, "qc_expr_dist_colon")
ggsave(paste0(f, ".png"), p, width=w, height=h)
ggsave(paste0(f, ".pdf"), p, width=w, height=h)

# plot background correction (before and after)
p <- stgeomx::plot_bgcorrect(fds_colon)
p
w <- 8
h <- 8
f <- file.path(correction_normalization_dir, "qc_bgcorrect_colon")
ggsave(paste0(f, ".png"), p, width=w, height=h)
ggsave(paste0(f, ".pdf"), p, width=w, height=h)

# background correction must use original unfiltered matrix
fds_colon <- stgeomx::bgcorrect(ds_colon, method=bgcorrect_method,
                          bw.adjust=bgcorrect_bw_adjust,
                          bg.quant=bgcorrect_bg_quant)
# apply filters before normalizing
fds_colon <- stgeomx::apply_filters(fds_colon)
# normalize
fds_colon <- stgeomx::normalize(fds_colon, norm_method)

# filtering stats
nrow(ds$samples)
nrow(fds_colon$samples)
table(ds$meta$keep)

# final normalized gene expression matrix
log2_norm_cpm_colon <- log2(fds_colon$x + 1)

# BRAIN
# background correction must use original unfiltered matrix
fds_brain <- stgeomx::bgcorrect(ds_brain, method=bgcorrect_method, 
                          bw.adjust=bgcorrect_bw_adjust, 
                          bg.quant=bgcorrect_bg_quant)
# normalize
fds_brain <- stgeomx::normalize(fds_brain, norm_method)

# plot expression distribution
p <- stgeomx::plot_expr_dist(fds_brain)
p
w <- 8
h <- 8
f <- file.path(correction_normalization_dir, "qc_expr_dist_brain")
ggsave(paste0(f, ".png"), p, width=w, height=h)
ggsave(paste0(f, ".pdf"), p, width=w, height=h)

# plot background correction (before and after)
p <- stgeomx::plot_bgcorrect(fds_brain)
p
w <- 8
h <- 8
f <- file.path(correction_normalization_dir, "qc_bgcorrect_brain")
ggsave(paste0(f, ".png"), p, width=w, height=h)
ggsave(paste0(f, ".pdf"), p, width=w, height=h)

# background correction must use original unfiltered matrix
fds_brain <- stgeomx::bgcorrect(ds_brain, method=bgcorrect_method,
                          bw.adjust=bgcorrect_bw_adjust,
                          bg.quant=bgcorrect_bg_quant)
# apply filters before normalizing
fds_brain <- stgeomx::apply_filters(fds_brain)
# normalize
fds_brain <- stgeomx::normalize(fds_brain, norm_method)

# filtering stats
nrow(ds$samples)
nrow(fds$samples)
table(ds$meta$keep)

# final normalized gene expression matrix
log2_norm_cpm_brain <- log2(fds_brain$x + 1)

```
# Manuscript figure generation

## Helper function that creates plots for figure 2 and 3
```{r helper, include=FALSE}
cpm <- function(x) {
  sweep(x, 2, colSums(x) / 1e6, FUN="/")
}

plot_bgcorrect_helper <- function(ds, quantiles=c(0.10, 0.25, 0.5, 0.75, 0.90, 0.95, 0.99)) {
  # gene metadata
  bg <- ds$meta$bg
  keep <- ds$meta$keep
  # labels for plot legend
  quant_names <- paste0("q", round(100 * quantiles))
  # sample columns used for plot
  s <- select(ds$samples, "aoi", "slide", "keep")
  # compute background level per aoi
  s$bg_cpm <- 1e6 * colSums(ds$counts[bg,]) / colSums(ds$counts)
 
  # filter genes
  x <- ds$x[(!bg) & keep,]
  counts <- ds$counts[(!bg) & keep,]
 
  # gene cpm before bgcorrect
  xgene <- as_tibble(cpm(counts))
  qgene <- xgene %>%
    reframe(across(everything(), ~ stats::quantile(.x, quantiles))) %>%
    mutate(name="gene", quant=quant_names)
 
  # cpm after bgcorrect
  y <- as_tibble(cpm(x))
  qy <- y %>%
    reframe(across(everything(), ~ stats::quantile(.x, quantiles))) %>%
    mutate(name="y", quant=quant_names)
 
  # combine bg and gene quantiles
  q <- bind_rows(qgene, qy) %>%
    tidyr::pivot_longer(colnames(x), names_to="aoi", values_to="value") %>%
    tidyr::pivot_wider(names_from="name", values_from="value", names_prefix="") %>%
    inner_join(s, by=c("aoi"="aoi"), suffix=c("_gene", "_aoi"))
 
  p <- ggplot(filter(q, .data$keep), aes(x=.data$bg_cpm+1, y=.data$y+1, color=.data$quant)) +
    geom_point(alpha=0.6) +
    geom_smooth(data=filter(q, .data$keep), linetype='dashed', method = 'lm',
                formula = y ~ x, se=FALSE, alpha=0.6) +
    scale_color_viridis_d(option="plasma") +
    theme_minimal() +
    labs(x="Noise (CPM)", y="Normalized Counts (CPM)")
  return(p)
}

most_var_genes <- function(x, ntop=500, span=0.50) {
  tbl <- bind_cols(
    u = rowMeans(x),
    v = apply(x, 1, function(x) var(x))
  )
  lovar <- loess(v ~ u, data=tbl, span=span)
  tbl$lovar <- predict(lovar)
  tbl$vadj <- tbl$v / tbl$lovar
  keep <- order(tbl$vadj, decreasing=TRUE)[seq_len(min(ntop, nrow(tbl)))]
  return(x[keep,])
}

pca <- function(s, y) {
  # PCA
  pca_res <- prcomp(t(y), center=TRUE, scale.=FALSE)
  pca_res.var <- round(100 * pca_res$sdev^2 / sum(pca_res$sdev^2 ), 2)
  pca_res$var_txt <- paste(colnames(pca_res$x), " (", paste(as.character(pca_res.var), "%", ")", sep=""), sep="")

  return(pca_res)
} 

plot_pca <- function(pca_res, x) {
  # plot dimensions
  width <- 12
  height <- 10
  
  # creation of individual PCA plots
  p <- ggplot(x, aes(x=PC1, y=PC2, color=study)) +
    geom_point(alpha = 0.4, size=1) +
    theme_bw() +
    theme(aspect.ratio=1) +
    theme(legend.text = element_text(size=7)) +
    labs(x=pca_res$var_txt[1], y=pca_res$var_txt[2])
  return(p)
}

plot_pca_bg <- function(pca_res, x) {
  # plot dimensions
  width <- 12
  height <- 10
  
  # creation of individual PCA plots
  p <- ggplot(x, aes(x=PC1, y=PC2, color=bg_auc)) +
    geom_point(alpha = 0.4, size=1) +
    scale_color_viridis_c() +
    theme_bw() +
    theme(aspect.ratio=1) +
    theme(legend.text = element_text(size=7)) +
    labs(x=pca_res$var_txt[1], y=pca_res$var_txt[2])
  return(p)
}

cluster_pca <- function(pca_res) {
# npcs
npcs <- 3
nclusters <- 5

# cluster PCA results
# pam
x <- pca_res$x[,1:npcs]
pamx <- cluster::pam(dist(x), k=nclusters)
memb <- paste0("c", pamx$clustering)

return(memb)
}

pca_plot_ipmn <- function(fds, pca_res, xvar, yvar, color_by, shape_by) {
  x <- bind_cols(fds$samples, as_tibble(pca_res$x))
  color_by_string <- rlang::as_string(rlang::ensym(color_by))
  shape_by_string <- rlang::as_string(rlang::ensym(shape_by))
  p <- ggplot(x, aes(x={{xvar}}, y={{yvar}}, color={{color_by}}, shape={{shape_by}})) +
    geom_point(alpha = 0.6, size=1) +
    # scale_color_manual(values = color_scales[[color_by_string]]) +
    # scale_shape_manual(values = shape_scales[[shape_by_string]]) +
    coord_fixed() +
    theme_bw()
  return(p)
}
```


## FIGURE 2: effects of bg correction and normalization on merged only
```{r bg correction and normalization methods plot}
library("patchwork")
correction_normalization_dir <- file.path(plot_dir, "bgcorrection_normalization")

# background correction must use original unfiltered matrix
fds_bgsub <- stgeomx::bgcorrect(ds, method="bgsub",
                          bw.adjust=bgcorrect_bw_adjust,
                          bg.quant=bgcorrect_bg_quant)
fds_none <- stgeomx::bgcorrect(ds, method="none",
                          bw.adjust=bgcorrect_bw_adjust,
                          bg.quant=bgcorrect_bg_quant)
fds_qq <- stgeomx::bgcorrect(ds, method="qq",
                          bw.adjust=bgcorrect_bw_adjust,
                          bg.quant=bgcorrect_bg_quant)

# normalize
fds_bgsub_qn <- stgeomx::normalize(fds_bgsub, "qn")
fds_bgsub_rle <- stgeomx::normalize(fds_bgsub, "rle")
fds_bgsub_q3 <- stgeomx::normalize(fds_bgsub, "q3")
fds_bgsub_tmm <- stgeomx::normalize(fds_bgsub, "tmm")

fds_qq_qn <- stgeomx::normalize(fds_qq, "qn")
fds_qq_rle <- stgeomx::normalize(fds_qq, "rle")
fds_qq_q3 <- stgeomx::normalize(fds_qq, "q3")
fds_qq_tmm <- stgeomx::normalize(fds_qq, "tmm")

fds_none_qn <- stgeomx::normalize(fds_none, "qn")
fds_none_rle <- stgeomx::normalize(fds_none, "rle")
fds_none_q3 <- stgeomx::normalize(fds_none, "q3")
fds_none_tmm <- stgeomx::normalize(fds_none, "tmm")

# plot background correction
plot_bgsub_qn <- stgeomx::plot_bgcorrect(fds_bgsub_qn)
plot_bgsub_rle <- stgeomx::plot_bgcorrect(fds_bgsub_rle)
plot_bgsub_q3 <- stgeomx::plot_bgcorrect(fds_bgsub_q3)
plot_bgsub_tmm <- stgeomx::plot_bgcorrect(fds_bgsub_tmm)

plot_qq_qn <- stgeomx::plot_bgcorrect(fds_qq_qn)
plot_qq_rle <- stgeomx::plot_bgcorrect(fds_qq_rle)
plot_qq_q3 <- stgeomx::plot_bgcorrect(fds_qq_q3)
plot_qq_tmm <- stgeomx::plot_bgcorrect(fds_qq_tmm)

plot_none_qn <- stgeomx::plot_bgcorrect(fds_none_qn)
plot_none_rle <- stgeomx::plot_bgcorrect(fds_none_rle)
plot_none_q3 <- stgeomx::plot_bgcorrect(fds_none_q3)
plot_none_tmm <- stgeomx::plot_bgcorrect(fds_none_tmm)

# Column and row titles
col_titles <- list(
  wrap_elements(grid::textGrob('', gp = grid::gpar(fontsize = 18, fontface = "bold"))),
  wrap_elements(grid::textGrob('Q3', gp = grid::gpar(fontsize = 18, fontface = "bold"))),
  wrap_elements(grid::textGrob('TMM', gp = grid::gpar(fontsize = 18, fontface = "bold"))),
  wrap_elements(grid::textGrob('RLE', gp = grid::gpar(fontsize = 18, fontface = "bold"))),
  wrap_elements(grid::textGrob('QN', gp = grid::gpar(fontsize = 18, fontface = "bold")))
)

row_titles <- list(
  wrap_elements(grid::textGrob('NONE', gp = grid::gpar(fontsize = 18, fontface = "bold"))),
  wrap_elements(grid::textGrob('BGSub', gp = grid::gpar(fontsize = 18, fontface = "bold"))),
  wrap_elements(grid::textGrob('QQ', gp = grid::gpar(fontsize = 18, fontface = "bold")))
)

# Create grid
p <- wrap_plots(
  col_titles[[1]], col_titles[[2]], col_titles[[3]], col_titles[[4]], col_titles[[5]],
  row_titles[[1]], plot_none_q3, plot_none_tmm, plot_none_rle, plot_none_qn,
  row_titles[[2]], plot_bgsub_q3, plot_bgsub_tmm, plot_bgsub_rle, plot_bgsub_qn,
  row_titles[[3]], plot_qq_q3, plot_qq_tmm, plot_qq_rle, plot_qq_qn,
  ncol = 5,
  heights = c(0.1, 1, 1, 1),
  widths = c(0.3, 1, 1, 1, 1)
) +
plot_layout(
  guides = 'collect'
) &
theme(plot.margin = unit(c(0.5, 1, 1, 1), "cm"))

# Add titles
p <- p +
  plot_annotation(
    title = "Effect of Background Correction and Normalization Methods on Normalized Expression Levels"
  ) &
  theme(
    plot.title = element_text(size = 22, face = "bold", hjust = 0.5)
  )

f <- file.path(correction_normalization_dir, "bgcorrect_normalization_grid")
ggsave(paste0(f, ".png"), p, width=30, height=15)
ggsave(paste0(f, ".pdf"), p, width=30, height=15)
```

## Setup for downstream analysis
```{r further processing}
# background correction must use original unfiltered matrix
fds_none <- stgeomx::bgcorrect(ds, method="none",
                          bw.adjust=bgcorrect_bw_adjust,
                          bg.quant=bgcorrect_bg_quant)
fds_bgsub <- stgeomx::bgcorrect(ds, method="bgsub",
                          bw.adjust=bgcorrect_bw_adjust,
                          bg.quant=bgcorrect_bg_quant)
fds_qq <- stgeomx::bgcorrect(ds, method="qq",
                          bw.adjust=bgcorrect_bw_adjust,
                          bg.quant=bgcorrect_bg_quant)

# apply filters before normalizing
fds_none <- stgeomx::apply_filters(fds_none)
fds_bgsub <- stgeomx::apply_filters(fds_bgsub)
fds_qq <- stgeomx::apply_filters(fds_qq)

# normalize
fds_none_q3 <- stgeomx::normalize(fds_none, "q3")
fds_none_tmm <- stgeomx::normalize(fds_none, "tmm")
fds_none_rle <- stgeomx::normalize(fds_none, "rle")
fds_none_qn <- stgeomx::normalize(fds_none, "qn")

fds_bgsub_q3 <- stgeomx::normalize(fds_bgsub, "q3")
fds_bgsub_tmm <- stgeomx::normalize(fds_bgsub, "tmm")
fds_bgsub_rle <- stgeomx::normalize(fds_bgsub, "rle")
fds_bgsub_qn <- stgeomx::normalize(fds_bgsub, "qn")

fds_qq_q3 <- stgeomx::normalize(fds_qq, "q3")
fds_qq_tmm <- stgeomx::normalize(fds_qq, "tmm")
fds_qq_rle <- stgeomx::normalize(fds_qq, "rle")
fds_qq_qn <- stgeomx::normalize(fds_qq, "qn")

# final normalized gene expression matrix
log2_norm_cpm_none_q3 <- log2(fds_none_q3$x + 1)
log2_norm_cpm_none_tmm <- log2(fds_none_tmm$x + 1)
log2_norm_cpm_none_rle <- log2(fds_none_rle$x + 1)
log2_norm_cpm_none_qn <- log2(fds_none_qn$x + 1)

log2_norm_cpm_bgsub_q3 <- log2(fds_bgsub_q3$x + 1)
log2_norm_cpm_bgsub_tmm <- log2(fds_bgsub_tmm$x + 1)
log2_norm_cpm_bgsub_rle <- log2(fds_bgsub_rle$x + 1)
log2_norm_cpm_bgsub_qn <- log2(fds_bgsub_qn$x + 1)

log2_norm_cpm_qq_q3 <- log2(fds_qq_q3$x + 1)
log2_norm_cpm_qq_tmm <- log2(fds_qq_tmm$x + 1)
log2_norm_cpm_qq_rle <- log2(fds_qq_rle$x + 1)
log2_norm_cpm_qq_qn <- log2(fds_qq_qn$x + 1)
```

## FIGURE 3: PCA grid
``` {r pca}
pca_dir <- file.path(plot_dir, "pca")

# subset most variable genes
ntop <- 500
myspan <- 0.5

# most var genes
s_none_q3 <- fds_none_q3$samples
y_none_q3 <- most_var_genes(log2_norm_cpm_none_q3, ntop=ntop, span=myspan)
s_none_tmm <- fds_none_tmm$samples
y_none_tmm <- most_var_genes(log2_norm_cpm_none_tmm, ntop=ntop, span=myspan)
s_none_rle <- fds_none_rle$samples
y_none_rle <- most_var_genes(log2_norm_cpm_none_rle, ntop=ntop, span=myspan)
s_none_qn <- fds_none_qn$samples
y_none_qn <- most_var_genes(log2_norm_cpm_none_qn, ntop=ntop, span=myspan)

s_bgsub_q3 <- fds_bgsub_q3$samples
y_bgsub_q3 <- most_var_genes(log2_norm_cpm_bgsub_q3, ntop=ntop, span=myspan)
s_bgsub_tmm <- fds_bgsub_tmm$samples
y_bgsub_tmm <- most_var_genes(log2_norm_cpm_bgsub_tmm, ntop=ntop, span=myspan)
s_bgsub_rle <- fds_bgsub_rle$samples
y_bgsub_rle <- most_var_genes(log2_norm_cpm_bgsub_rle, ntop=ntop, span=myspan)
s_bgsub_qn <- fds_bgsub_qn$samples
y_bgsub_qn <- most_var_genes(log2_norm_cpm_bgsub_qn, ntop=ntop, span=myspan)

s_qq_q3 <- fds_qq_q3$samples
y_qq_q3 <- most_var_genes(log2_norm_cpm_qq_q3, ntop=ntop, span=myspan)
s_qq_tmm <- fds_qq_tmm$samples
y_qq_tmm <- most_var_genes(log2_norm_cpm_qq_tmm, ntop=ntop, span=myspan)
s_qq_rle <- fds_qq_rle$samples
y_qq_rle <- most_var_genes(log2_norm_cpm_qq_rle, ntop=ntop, span=myspan)
s_qq_qn <- fds_qq_qn$samples
y_qq_qn <- most_var_genes(log2_norm_cpm_qq_qn, ntop=ntop, span=myspan)

# PCA_res
pca_res_none_q3 <- pca(s_none_q3, y_none_q3)
pca_res_none_tmm <- pca(s_none_tmm, y_none_tmm)
pca_res_none_rle <- pca(s_none_rle, y_none_rle)
pca_res_none_qn <- pca(s_none_qn, y_none_qn)

pca_res_bgsub_q3 <- pca(s_bgsub_q3, y_bgsub_q3)
pca_res_bgsub_tmm <- pca(s_bgsub_tmm, y_bgsub_tmm)
pca_res_bgsub_rle <- pca(s_bgsub_rle, y_bgsub_rle)
pca_res_bgsub_qn <- pca(s_bgsub_qn, y_bgsub_qn)

pca_res_qq_q3 <- pca(s_qq_q3, y_qq_q3)
pca_res_qq_tmm <- pca(s_qq_tmm, y_qq_tmm)
pca_res_qq_rle <- pca(s_qq_rle, y_qq_rle)
pca_res_qq_qn <- pca(s_qq_qn, y_qq_qn)

# x
x_none_q3 <- bind_cols(s_none_q3, as_tibble(pca_res_none_q3$x))
x_none_tmm <- bind_cols(s_none_tmm, as_tibble(pca_res_none_tmm$x))
x_none_rle <- bind_cols(s_none_rle, as_tibble(pca_res_none_rle$x))
x_none_qn <- bind_cols(s_none_qn, as_tibble(pca_res_none_qn$x))

x_bgsub_q3 <- bind_cols(s_bgsub_q3, as_tibble(pca_res_bgsub_q3$x))
x_bgsub_tmm <- bind_cols(s_bgsub_tmm, as_tibble(pca_res_bgsub_tmm$x))
x_bgsub_rle <- bind_cols(s_bgsub_rle, as_tibble(pca_res_bgsub_rle$x))
x_bgsub_qn <- bind_cols(s_bgsub_qn, as_tibble(pca_res_bgsub_qn$x))

x_qq_q3 <- bind_cols(s_qq_q3, as_tibble(pca_res_qq_q3$x))
x_qq_tmm <- bind_cols(s_qq_tmm, as_tibble(pca_res_qq_tmm$x))
x_qq_rle <- bind_cols(s_qq_rle, as_tibble(pca_res_qq_rle$x))
x_qq_qn <- bind_cols(s_qq_qn, as_tibble(pca_res_qq_qn$x))

# individual pca plots
pca_none_q3 <- plot_pca(pca_res_none_q3, x_none_q3)
pca_none_tmm <- plot_pca(pca_res_none_tmm, x_none_tmm)
pca_none_rle <- plot_pca(pca_res_none_q3, x_none_rle)
pca_none_qn <- plot_pca(pca_res_none_q3, x_none_qn)

pca_bgsub_q3 <- plot_pca(pca_res_bgsub_q3, x_bgsub_q3)
pca_bgsub_tmm <- plot_pca(pca_res_bgsub_tmm, x_bgsub_tmm)
pca_bgsub_rle <- plot_pca(pca_res_bgsub_q3, x_bgsub_rle)
pca_bgsub_qn <- plot_pca(pca_res_bgsub_q3, x_bgsub_qn)

pca_qq_q3 <- plot_pca(pca_res_qq_q3, x_qq_q3)
pca_qq_tmm <- plot_pca(pca_res_qq_tmm, x_qq_tmm)
pca_qq_rle <- plot_pca(pca_res_qq_q3, x_qq_rle)
pca_qq_qn <- plot_pca(pca_res_qq_q3, x_qq_qn)


# Grid column and row titles
col_titles <- list(
  wrap_elements(grid::textGrob('', gp = grid::gpar(fontsize = 18, fontface = "bold"))),
  wrap_elements(grid::textGrob('Q3', gp = grid::gpar(fontsize = 18, fontface = "bold"))),
  wrap_elements(grid::textGrob('TMM', gp = grid::gpar(fontsize = 18, fontface = "bold"))),
  wrap_elements(grid::textGrob('RLE', gp = grid::gpar(fontsize = 18, fontface = "bold"))),
  wrap_elements(grid::textGrob('QN', gp = grid::gpar(fontsize = 18, fontface = "bold")))
)

row_titles <- list(
  wrap_elements(grid::textGrob('NONE', gp = grid::gpar(fontsize = 18, fontface = "bold"))),
  wrap_elements(grid::textGrob('BGSub', gp = grid::gpar(fontsize = 18, fontface = "bold"))),
  wrap_elements(grid::textGrob('QQ', gp = grid::gpar(fontsize = 18, fontface = "bold")))
)

# Create grid
p <- wrap_plots(
  col_titles[[1]], col_titles[[2]], col_titles[[3]], col_titles[[4]], col_titles[[5]],
  row_titles[[1]], pca_none_q3, pca_none_tmm, pca_none_rle, pca_none_qn,
  row_titles[[2]], pca_bgsub_q3, pca_bgsub_tmm, pca_bgsub_rle, pca_bgsub_qn,
  row_titles[[3]], pca_qq_q3, pca_qq_tmm, pca_qq_rle, pca_qq_qn,
  ncol = 5,
  heights = c(0.1, 1, 1, 1),
  widths = c(0.3, 1, 1, 1, 1)
) +
plot_layout(
  guides = 'collect'
) &
theme(plot.margin = unit(c(0.5, 1, 1, 1), "cm"))

# Add titles
p <- p + 
  plot_annotation(
    title = "Effect of Background Correction and Normalization Methods on PCA"
  ) & 
  theme(
    plot.title = element_text(size = 22, face = "bold", hjust = 0.5)
  )

f <- file.path(pca_dir, "pca_grid_500")
ggsave(paste0(f, ".png"), p, width=20, height=15)
ggsave(paste0(f, ".pdf"), p, width=20, height=15)

```

### FIGURE 3B: Background Noise
``` {r}
# individual pca background plots
pca_none_q3 <- plot_pca_bg(pca_res_none_q3, x_none_q3)
pca_none_tmm <- plot_pca_bg(pca_res_none_tmm, x_none_tmm)
pca_none_rle <- plot_pca_bg(pca_res_none_q3, x_none_rle)
pca_none_qn <- plot_pca_bg(pca_res_none_q3, x_none_qn)

pca_bgsub_q3 <- plot_pca_bg(pca_res_bgsub_q3, x_bgsub_q3)
pca_bgsub_tmm <- plot_pca_bg(pca_res_bgsub_tmm, x_bgsub_tmm)
pca_bgsub_rle <- plot_pca_bg(pca_res_bgsub_q3, x_bgsub_rle)
pca_bgsub_qn <- plot_pca_bg(pca_res_bgsub_q3, x_bgsub_qn)

pca_qq_q3 <- plot_pca_bg(pca_res_qq_q3, x_qq_q3)
pca_qq_tmm <- plot_pca_bg(pca_res_qq_tmm, x_qq_tmm)
pca_qq_rle <- plot_pca_bg(pca_res_qq_q3, x_qq_rle)
pca_qq_qn <- plot_pca_bg(pca_res_qq_q3, x_qq_qn)


# Grid column and row titles
col_titles <- list(
  wrap_elements(grid::textGrob('', gp = grid::gpar(fontsize = 18, fontface = "bold"))),
  wrap_elements(grid::textGrob('Q3', gp = grid::gpar(fontsize = 18, fontface = "bold"))),
  wrap_elements(grid::textGrob('TMM', gp = grid::gpar(fontsize = 18, fontface = "bold"))),
  wrap_elements(grid::textGrob('RLE', gp = grid::gpar(fontsize = 18, fontface = "bold"))),
  wrap_elements(grid::textGrob('QN', gp = grid::gpar(fontsize = 18, fontface = "bold")))
)

row_titles <- list(
  wrap_elements(grid::textGrob('NONE', gp = grid::gpar(fontsize = 18, fontface = "bold"))),
  wrap_elements(grid::textGrob('BGSub', gp = grid::gpar(fontsize = 18, fontface = "bold"))),
  wrap_elements(grid::textGrob('QQ', gp = grid::gpar(fontsize = 18, fontface = "bold")))
)

# Create grid
p <- wrap_plots(
  col_titles[[1]], col_titles[[2]], col_titles[[3]], col_titles[[4]], col_titles[[5]],
  row_titles[[1]], pca_none_q3, pca_none_tmm, pca_none_rle, pca_none_qn,
  row_titles[[2]], pca_bgsub_q3, pca_bgsub_tmm, pca_bgsub_rle, pca_bgsub_qn,
  row_titles[[3]], pca_qq_q3, pca_qq_tmm, pca_qq_rle, pca_qq_qn,
  ncol = 5,
  heights = c(0.1, 1, 1, 1),
  widths = c(0.3, 1, 1, 1, 1)
) +
plot_layout(
  guides = 'collect'
) &
theme(plot.margin = unit(c(0.5, 1, 1, 1), "cm"))

# Add titles
p <- p + 
  plot_annotation(
    title = "Effect of Background Correction and Normalization Methods on PCA"
  ) & 
  theme(
    plot.title = element_text(size = 22, face = "bold", hjust = 0.5)
  )

# Save graph
f <- file.path(pca_dir, "pca_grid_bg")
ggsave(paste0(f, ".png"), p, width=20, height=15)
ggsave(paste0(f, ".pdf"), p, width=20, height=15)
```

### Pearson correlates
```{r}

methods <- c("none_q3", "none_tmm", "none_rle", "none_qn", 
             "bgsub_q3", "bgsub_tmm", "bgsub_rle", "bgsub_qn", 
             "qq_q3", "qq_tmm", "qq_rle", "qq_qn")

# Performing cor.test and storing the result in lists
pc1_data <- list(cor.test(x_none_q3$PC1, x_none_q3$bg_auc),
                 cor.test(x_none_tmm$PC1, x_none_tmm$bg_auc),
                 cor.test(x_none_rle$PC1, x_none_rle$bg_auc),
                 cor.test(x_none_qn$PC1, x_none_qn$bg_auc),
                 cor.test(x_bgsub_q3$PC1, x_bgsub_q3$bg_auc),
                 cor.test(x_bgsub_tmm$PC1, x_bgsub_tmm$bg_auc),
                 cor.test(x_bgsub_rle$PC1, x_bgsub_rle$bg_auc),
                 cor.test(x_bgsub_qn$PC1, x_bgsub_qn$bg_auc),
                 cor.test(x_qq_q3$PC1, x_qq_q3$bg_auc),
                 cor.test(x_qq_tmm$PC1, x_qq_tmm$bg_auc),
                 cor.test(x_qq_rle$PC1, x_qq_rle$bg_auc),
                 cor.test(x_qq_qn$PC1, x_qq_qn$bg_auc))

pc2_data <- list(cor.test(x_none_q3$PC2, x_none_q3$bg_auc),
                 cor.test(x_none_tmm$PC2, x_none_tmm$bg_auc),
                 cor.test(x_none_rle$PC2, x_none_rle$bg_auc),
                 cor.test(x_none_qn$PC2, x_none_qn$bg_auc),
                 cor.test(x_bgsub_q3$PC2, x_bgsub_q3$bg_auc),
                 cor.test(x_bgsub_tmm$PC2, x_bgsub_tmm$bg_auc),
                 cor.test(x_bgsub_rle$PC2, x_bgsub_rle$bg_auc),
                 cor.test(x_bgsub_qn$PC2, x_bgsub_qn$bg_auc),
                 cor.test(x_qq_q3$PC2, x_qq_q3$bg_auc),
                 cor.test(x_qq_tmm$PC2, x_qq_tmm$bg_auc),
                 cor.test(x_qq_rle$PC2, x_qq_rle$bg_auc),
                 cor.test(x_qq_qn$PC2, x_qq_qn$bg_auc))

pc3_data <- list(cor.test(x_none_q3$PC3, x_none_q3$bg_auc),
                 cor.test(x_none_tmm$PC3, x_none_tmm$bg_auc),
                 cor.test(x_none_rle$PC3, x_none_rle$bg_auc),
                 cor.test(x_none_qn$PC3, x_none_qn$bg_auc),
                 cor.test(x_bgsub_q3$PC3, x_bgsub_q3$bg_auc),
                 cor.test(x_bgsub_tmm$PC3, x_bgsub_tmm$bg_auc),
                 cor.test(x_bgsub_rle$PC3, x_bgsub_rle$bg_auc),
                 cor.test(x_bgsub_qn$PC3, x_bgsub_qn$bg_auc),
                 cor.test(x_qq_q3$PC3, x_qq_q3$bg_auc),
                 cor.test(x_qq_tmm$PC3, x_qq_tmm$bg_auc),
                 cor.test(x_qq_rle$PC3, x_qq_rle$bg_auc),
                 cor.test(x_qq_qn$PC3, x_qq_qn$bg_auc))

# Extract the correlation estimates from cor.test results
pc1_estimates <- sapply(pc1_data, function(test) test$estimate)
pc2_estimates <- sapply(pc2_data, function(test) test$estimate)
pc3_estimates <- sapply(pc3_data, function(test) test$estimate)

# Initialize an empty matrix to store the results
results <- matrix(NA, nrow = length(methods), ncol = 3)
rownames(results) <- methods
colnames(results) <- c("PC1", "PC2", "PC3")

# Populate the results matrix
results[, "PC1"] <- pc1_estimates
results[, "PC2"] <- pc2_estimates
results[, "PC3"] <- pc3_estimates

# Convert to a data frame for better readability
results_df <- as.data.frame(results)

# Print the results
print(results_df)
```

### PC1, PC2, PC3, PC4, PC5 for qq_qn
``` {r pcs}
s <- fds_qq_qn$samples
pcs <- c("PC1", "PC2", "PC3", "PC4")
pca_tbl <- pca_res_qq_qn$x
tbl <- as_tibble(pca_tbl[,pcs], rownames="aoi")
 
x <- tbl %>% pivot_longer(cols=all_of(pcs), names_to="xpc", values_to="xvalue")
y <- tbl %>% pivot_longer(cols=all_of(pcs), names_to="ypc", values_to="yvalue")
tbl <- full_join(x, y, by=join_by(aoi), relationship="many-to-many")
tbl <- left_join(tbl, s, by=join_by(aoi))
 
custom_theme_pca <- function() {
  theme_minimal() +
  theme(aspect.ratio=1) +
    theme(legend.title = element_text(size=8),
          legend.text = element_text(size=7),
          legend.position = "bottom")
}
 
p <- ggplot(filter(tbl, xpc < ypc), aes(x=xvalue, y=yvalue, color=region, shape=study)) +
  geom_point(alpha=0.7) +
  custom_theme_pca() +
  facet_grid(ypc ~ xpc)
 
f <- file.path(pca_dir, "pca_1234_grid")
ggsave(paste0(f, ".png"), p, width=25, height=20)
ggsave(paste0(f, ".pdf"), p, width=25, height=20)
```

### SUPPLEMENTAL: regions on PCA
``` {r PCA regions}
# pancreas region
p <- ggplot(x_qq_qn %>% arrange(study), aes(x=PC1, y=PC2, color = ifelse(study == "pancreas", region, "other"), shape=study)) +
  geom_point(alpha = 0.8, size=3) +
  scale_color_manual(values = c("other" = "grey",
                                "acini" = "red",
                                "acini_and_duct" = "orange",
                                "alpha_cell_enriched_islet" = "green",
                                "beta_cell_enriched_islet" = "blue",
                                "duct" = "yellow",
                                "islet" = "pink")) +
  theme_bw() +
  theme(aspect.ratio=1) +
  theme(legend.text = element_text(size=7)) +
  labs(x=pca_res_qq_qn$var_txt[1], y=pca_res_qq_qn$var_txt[2])
f <- file.path(plot_dir, paste0("pca_region_pancreas.pdf"))
ggsave(f, plot=p, device="pdf", width=10, height=10)

# lymph node and colon regions
p <- ggplot(x_qq_qn, aes(x = PC1, y = PC2, color = ifelse(study %in% c("colon", "ln"), region, "other"), shape = study)) +
  geom_point(alpha = 0.8, size=3) +
  scale_color_manual(values = c("other" = "grey",
                                "b_cell_zone" = "red",
                                "circular_muscle_layer" = "salmon",
                                "connective_tissues" = "orange",
                                "enteroendocrine_cells" = "yellow",
                                "epithelium" = "green",
                                "germinal_center" = "palegreen3",
                                "lamina_propria" = "cyan4",
                                "longitudinal_muscle_layer" = "turquoise1",
                                "lymphoid_structure" = "steelblue",
                                "medulla" = "blue",
                                "muscularis_mucosae" = "darkmagenta",
                                "paneth_cells" = "purple",
                                "t_cell_zone" = "pink",
                                "trabecula" = "deeppink",
                                "vessels" = "tan3")) +
  theme_bw() +
  theme(aspect.ratio=1) +
  theme(legend.text = element_text(size=7)) +
  labs(x=pca_res_qq_qn$var_txt[1], y=pca_res_qq_qn$var_txt[2])
p
f <- file.path(plot_dir, paste0("pca_region_ln_colon.pdf"))
ggsave(f, plot=p, device="pdf", width=20, height=15)
```

## FIGURE 4: Clustering
``` {r clustering}
# the following guides were used in this section:
# guide: https://rpubs.com/KarolinaSzczesna/862710
# guide: https://rpubs.com/Bury/ClusteringOnPcaResults

# cluster, then add cluster membership to sample properties
fds_none_q3$samples$cluster <- cluster_pca(pca_res_none_q3)
fds_none_tmm$samples$cluster <- cluster_pca(pca_res_none_tmm)
fds_none_rle$samples$cluster <- cluster_pca(pca_res_none_rle)
fds_none_qn$samples$cluster <- cluster_pca(pca_res_none_qn)

fds_bgsub_q3$samples$cluster <- cluster_pca(pca_res_bgsub_q3)
fds_bgsub_tmm$samples$cluster <- cluster_pca(pca_res_bgsub_tmm)
fds_bgsub_rle$samples$cluster <- cluster_pca(pca_res_bgsub_rle)
fds_bgsub_qn$samples$cluster <- cluster_pca(pca_res_bgsub_qn)

fds_qq_q3$samples$cluster <- cluster_pca(pca_res_qq_q3)
fds_qq_tmm$samples$cluster <- cluster_pca(pca_res_qq_tmm)
fds_qq_rle$samples$cluster <- cluster_pca(pca_res_qq_rle)
fds_qq_qn$samples$cluster <- cluster_pca(pca_res_qq_qn)

# pca plot clustering
cluster_none_q3 <- pca_plot_ipmn(fds_none_q3, pca_res_none_q3, PC1, PC2, cluster, study)
cluster_none_tmm <- pca_plot_ipmn(fds_none_tmm, pca_res_none_tmm, PC1, PC2, cluster, study)
cluster_none_rle <- pca_plot_ipmn(fds_none_rle, pca_res_none_rle, PC1, PC2, cluster, study)
cluster_none_qn <- pca_plot_ipmn(fds_none_qn, pca_res_none_qn, PC1, PC2, cluster, study)

cluster_bgsub_q3 <- pca_plot_ipmn(fds_bgsub_q3, pca_res_bgsub_q3, PC1, PC2, cluster, study)
cluster_bgsub_tmm <- pca_plot_ipmn(fds_bgsub_tmm, pca_res_bgsub_tmm, PC1, PC2, cluster, study)
cluster_bgsub_rle <- pca_plot_ipmn(fds_bgsub_rle, pca_res_bgsub_rle, PC1, PC2, cluster, study)
cluster_bgsub_qn <- pca_plot_ipmn(fds_bgsub_qn, pca_res_bgsub_qn, PC1, PC2, cluster, study)

cluster_qq_q3 <- pca_plot_ipmn(fds_qq_q3, pca_res_qq_q3, PC1, PC2, cluster, study)
cluster_qq_tmm <- pca_plot_ipmn(fds_qq_tmm, pca_res_qq_tmm, PC1, PC2, cluster, study)
cluster_qq_rle <- pca_plot_ipmn(fds_qq_rle, pca_res_qq_rle, PC1, PC2, cluster, study)
cluster_qq_qn <- pca_plot_ipmn(fds_qq_qn, pca_res_qq_qn, PC1, PC2, cluster, study)

# Grid column and row titles
col_titles <- list(
  wrap_elements(grid::textGrob('', gp = grid::gpar(fontsize = 18, fontface = "bold"))),
  wrap_elements(grid::textGrob('Q3', gp = grid::gpar(fontsize = 18, fontface = "bold"))),
  wrap_elements(grid::textGrob('TMM', gp = grid::gpar(fontsize = 18, fontface = "bold"))),
  wrap_elements(grid::textGrob('RLE', gp = grid::gpar(fontsize = 18, fontface = "bold"))),
  wrap_elements(grid::textGrob('QN', gp = grid::gpar(fontsize = 18, fontface = "bold")))
)

row_titles <- list(
  wrap_elements(grid::textGrob('NONE', gp = grid::gpar(fontsize = 18, fontface = "bold"))),
  wrap_elements(grid::textGrob('BGSub', gp = grid::gpar(fontsize = 18, fontface = "bold"))),
  wrap_elements(grid::textGrob('QQ', gp = grid::gpar(fontsize = 18, fontface = "bold")))
)

# Create grid
p <- wrap_plots(
  col_titles[[1]], col_titles[[2]], col_titles[[3]], col_titles[[4]], col_titles[[5]],
  row_titles[[1]], cluster_none_q3, cluster_none_tmm, cluster_none_rle, cluster_none_qn,
  row_titles[[2]], cluster_bgsub_q3, cluster_bgsub_tmm, cluster_bgsub_rle, cluster_bgsub_qn,
  row_titles[[3]], cluster_qq_q3, cluster_qq_tmm, cluster_qq_rle, cluster_qq_qn,
  ncol = 5,
  heights = c(0.1, 1, 1, 1),
  widths = c(0.3, 1, 1, 1, 1)
) +
plot_layout(
  guides = 'collect'
) &
theme(plot.margin = unit(c(0.5, 1, 1, 1), "cm"))

# Add titles
p <- p + 
  plot_annotation(
    title = "Effect of Background Correction and Normalization Methods on PCA Clustering"
  ) & 
  theme(
    plot.title = element_text(size = 22, face = "bold", hjust = 0.5)
  )

f <- file.path(pca_dir, "pca_clustering_grid")
ggsave(paste0(f, ".png"), p, width=20, height=15)
ggsave(paste0(f, ".pdf"), p, width=20, height=15)
```

### Optimal clustering
``` {r optimal clustering}
npcs <- 3

# assess optimal number of clusters
x <- pca_res_qq_qn$x[,1:npcs]

p <- fviz_nbclust(x, FUNcluster=cluster::pam, method="wss")
f <- file.path(pca_dir, "pca_fviz_nbclust_wss")
ggsave(paste0(f, ".png"), p, width=4, height=4)
```

### Assign names to clusters
``` {r}

fds_none_q3$samples <- fds_none_q3$samples %>% mutate(
  cluster_name = case_when(cluster == "c1" ~ "pancreas",
                           cluster == "c2" ~ "ln",
                           cluster == "c3" ~ "colon",
                           cluster == "c4" ~ "brain",
                           cluster == "c5" ~ "kidney")
)
x <- bind_cols(fds_none_q3$samples, as_tibble(pca_res_none_q3$x))

fds_none_tmm$samples <- fds_none_tmm$samples %>% mutate(
  cluster_name = case_when(cluster == "c1" ~ "pancreas",
                           cluster == "c2" ~ "colon",
                           cluster == "c3" ~ "ln",
                           cluster == "c4" ~ "brain",
                           cluster == "c5" ~ "kidney")
)
x <- bind_cols(fds_none_tmm$samples, as_tibble(pca_res_none_tmm$x))

fds_none_rle$samples <- fds_none_rle$samples %>% mutate(
  cluster_name = case_when(cluster == "c1" ~ "pancreas",
                           cluster == "c2" ~ "colon",
                           cluster == "c3" ~ "ln",
                           cluster == "c4" ~ "brain",
                           cluster == "c5" ~ "kidney")
)
x <- bind_cols(fds_none_rle$samples, as_tibble(pca_res_none_rle$x))

fds_none_qn$samples <- fds_none_qn$samples %>% mutate(
  cluster_name = case_when(cluster == "c1" ~ "pancreas",
                           cluster == "c2" ~ "ln",
                           cluster == "c3" ~ "colon",
                           cluster == "c4" ~ "brain",
                           cluster == "c5" ~ "kidney")
)
x <- bind_cols(fds_none_qn$samples, as_tibble(pca_res_none_qn$x))


fds_bgsub_q3$samples <- fds_bgsub_q3$samples %>% mutate(
  cluster_name = case_when(cluster == "c1" ~ "pancreas",
                           cluster == "c2" ~ "ln",
                           cluster == "c3" ~ "colon",
                           cluster == "c4" ~ "kidney",
                           cluster == "c5" ~ "brain")
)
x <- bind_cols(fds_bgsub_q3$samples, as_tibble(pca_res_bgsub_q3$x))

fds_bgsub_tmm$samples <- fds_bgsub_tmm$samples %>% mutate(
  cluster_name = case_when(cluster == "c1" ~ "pancreas",
                           cluster == "c2" ~ "ln",
                           cluster == "c3" ~ "colon",
                           cluster == "c4" ~ "kidney",
                           cluster == "c5" ~ "brain")
)
x <- bind_cols(fds_bgsub_tmm$samples, as_tibble(pca_res_bgsub_tmm$x))

fds_bgsub_rle$samples <- fds_bgsub_rle$samples %>% mutate(
  cluster_name = case_when(cluster == "c1" ~ "pancreas",
                           cluster == "c2" ~ "ln",
                           cluster == "c3" ~ "colon",
                           cluster == "c4" ~ "kidney",
                           cluster == "c5" ~ "brain")
)
x <- bind_cols(fds_bgsub_rle$samples, as_tibble(pca_res_bgsub_rle$x))

fds_bgsub_qn$samples <- fds_bgsub_qn$samples %>% mutate(
  cluster_name = case_when(cluster == "c1" ~ "pancreas",
                           cluster == "c2" ~ "ln",
                           cluster == "c3" ~ "colon",
                           cluster == "c4" ~ "kidney",
                           cluster == "c5" ~ "brain")
)
x <- bind_cols(fds_bgsub_qn$samples, as_tibble(pca_res_bgsub_qn$x))


fds_qq_q3$samples <- fds_qq_q3$samples %>% mutate(
  cluster_name = case_when(cluster == "c1" ~ "pancreas",
                           cluster == "c2" ~ "ln",
                           cluster == "c3" ~ "colon",
                           cluster == "c4" ~ "kidney",
                           cluster == "c5" ~ "brain")
)
x <- bind_cols(fds_qq_q3$samples, as_tibble(pca_res_qq_q3$x))

fds_qq_tmm$samples <- fds_qq_tmm$samples %>% mutate(
  cluster_name = case_when(cluster == "c1" ~ "pancreas",
                           cluster == "c2" ~ "ln",
                           cluster == "c3" ~ "colon",
                           cluster == "c4" ~ "brain",
                           cluster == "c5" ~ "kidney")
)
x <- bind_cols(fds_qq_tmm$samples, as_tibble(pca_res_qq_tmm$x))

fds_qq_rle$samples <- fds_qq_rle$samples %>% mutate(
  cluster_name = case_when(cluster == "c1" ~ "pancreas",
                           cluster == "c2" ~ "ln",
                           cluster == "c3" ~ "colon",
                           cluster == "c4" ~ "brain",
                           cluster == "c5" ~ "kidney")
)
x <- bind_cols(fds_qq_rle$samples, as_tibble(pca_res_qq_rle$x))

fds_qq_qn$samples <- fds_qq_qn$samples %>% mutate(
  cluster_name = case_when(cluster == "c1" ~ "pancreas",
                           cluster == "c2" ~ "ln",
                           cluster == "c3" ~ "colon",
                           cluster == "c4" ~ "kidney",
                           cluster == "c5" ~ "brain")
)
x <- bind_cols(fds_qq_qn$samples, as_tibble(pca_res_qq_qn$x))
```


### Cluster purity
```{r purity}
library("IntNMF")

# List of methods
methods <- list(fds_none_q3, fds_none_tmm, fds_none_rle, fds_none_qn, 
                fds_bgsub_q3, fds_bgsub_tmm, fds_bgsub_rle, fds_bgsub_qn, 
                fds_qq_q3, fds_qq_tmm, fds_qq_rle, fds_qq_qn)

# Initialize vectors to store Chi-squared statistics and p-values
purity_table <- numeric(length(methods))
method_names <- c("none_q3", "none_tmm", "none_rle", "none_qn",
                   "bgsub_q3", "bgsub_tmm", "bgsub_rle", "bgsub_qn",
                   "qq_q3", "qq_tmm", "qq_rle", "qq_qn")
purity_table <- NULL

# Loop through each method
for (i in methods) {

  # create purity score and add to purity table
  purity_table <- rbind(purity_table, ClusterPurity(i$samples$study, i$samples$cluster_name))
}

# Combine results into a data frame
results_df <- data.frame(
  Method = method_names,
  Purity_Scores = purity_table
 )

# Print the results
print(results_df)


```

### Chi-squared analysis of clusters
``` {r}
library(reshape2)

# List of methods
methods <- list(fds_none_q3, fds_none_tmm, fds_none_rle, fds_none_qn, 
                fds_bgsub_q3, fds_bgsub_tmm, fds_bgsub_rle, fds_bgsub_qn, 
                fds_qq_q3, fds_qq_tmm, fds_qq_rle, fds_qq_qn)

# Initialize vectors to store Chi-squared statistics and p-values
chi_squared_stats <- numeric(length(methods))
method_names <- c("none_q3", "none_tmm", "none_rle", "none_qn",
                   "bgsub_q3", "bgsub_tmm", "bgsub_rle", "bgsub_qn",
                   "qq_q3", "qq_tmm", "qq_rle", "qq_qn")
master_contingency_table <- NULL

# Loop through each method
for (i in seq_along(methods)) {
  # Create the contingency table
  contingency_table <- table(methods[[i]]$samples$study, methods[[i]]$samples$cluster_name)
  colnames(contingency_table) <- paste0(colnames(contingency_table), "_", method_names[i])

  
  # Print the contingency table
  print(contingency_table)
  
  # add to master contingency table
  if (i == 1) {
    master_contingency_table <- contingency_table
  }
  else {
    master_contingency_table <- cbind(master_contingency_table, contingency_table)
  }
  
  # Perform the Chi-squared test
  chi_squared_result <- chisq.test(contingency_table)
  print(chi_squared_result)
  
  # Store the Chi-squared statistic and p-value
  chi_squared_stats[i] <- chi_squared_result$statistic
}

# Combine results into a data frame
results_df <- data.frame(
  Method = method_names,
  Chi_Squared_Statistic = chi_squared_stats
 )

# Print the results
print(results_df)

print(master_contingency_table)
# heatmap(master_contingency_table)

# # Dummy data
# x <- LETTERS[1:20]
# y <- paste0("var", seq(1,20))
# data <- expand.grid(X=x, Y=y)
# data$Z <- runif(400, 0, 5)
 
# # Heatmap 
# master_contingency_table <- melt(master_contingency_table)
# ggplot(master_contingency_table, aes(x = Var1, y = Var2, fill = value)) +
# geom_tile()

```

# FIGURE 5: DE
## DE analysis

```{r}
library("limma")
library("writexl")

de_setup <- function(method_name, de_padj_cutoff=0.05, de_log2fc_cutoff=1) {
  # access fds and log2_norm_cpm
  fds_name <- paste0("fds_", method_name)
  fds <- get(fds_name)
  log2_name <- paste0("log2_norm_cpm_", method_name)
  log2_norm_cpm <- get(log2_name)
  
  # init de results
  de <- NULL
  
  # DE analysis of clusters
  s <- fds$samples
  y <- log2_norm_cpm[, s$aoi]
  s$cluster <- factor(s$cluster_name, levels=c("pancreas", "ln", "brain", "colon", "kidney"))
  
  # compare each cluster to all of the rest
  design <- model.matrix(~0 + cluster, s)
  contrasts <- makeContrasts(clusterPanc = clusterpancreas - (clusterln + clusterbrain + clusterkidney + clustercolon)/4,
                             clusterLn = clusterln - (clusterpancreas + clusterbrain + clusterkidney + clustercolon)/4,
                             clusterBrain = clusterbrain - (clusterln + clusterpancreas + clusterkidney + clustercolon)/4,
                             clusterKidney = clusterkidney - (clusterln + clusterbrain + clusterpancreas + clustercolon)/4,
                             clusterColon = clustercolon - (clusterln + clusterbrain + clusterkidney + clusterpancreas)/4,
                             levels = design)
  
  res <- run_limma_trend(y, design, contrasts,
                         padj_cutoff=de_padj_cutoff,
                         log2fc_cutoff=de_log2fc_cutoff)
  
  # write cluster results
  file_name <- paste0("de_clusters_", method_name, ".xlsx")
  write_xlsx(res, file.path(results_dir, file_name))
  de <- bind_rows(de, res)
  return(de)
}

run_limma_trend <- function(y, design, contrasts, padj_cutoff, log2fc_cutoff) {
  fit <- lmFit(y, design)
  fit <- contrasts.fit(fit, contrasts)
  fit <- eBayes(fit, trend=TRUE)

  x <- NULL
  for (coef in colnames(contrasts)) {
    res <- topTable(fit, coef=coef, number=Inf, sort.by="none")
    res <- as_tibble(res, rownames="gene")
    res <- res %>%
      select(gene,
             log2fc=logFC,
             avgexpr=AveExpr,
             pval=P.Value,
             padj=adj.P.Val) %>%
      mutate(de = case_when(padj > padj_cutoff ~ "no",
                            log2fc < -log2fc_cutoff ~ "dn",
                            log2fc > log2fc_cutoff ~ "up",
                            TRUE ~ "no"),
             contrast = coef)
    x <- bind_rows(x, res)
  }
  return(x)
}

de_none_q3 <- de_setup("none_q3")
de_none_tmm <- de_setup("none_tmm")
de_none_rle <- de_setup("none_rle")
de_none_qn <- de_setup("none_qn")

de_bgsub_q3 <- de_setup("bgsub_q3")
de_bgsub_tmm <- de_setup("bgsub_tmm")
de_bgsub_rle <- de_setup("bgsub_rle")
de_bgsub_qn <- de_setup("bgsub_qn")

de_qq_q3 <- de_setup("qq_q3")
de_qq_tmm <- de_setup("qq_tmm")
de_qq_rle <- de_setup("qq_rle")
de_qq_qn <- de_setup("qq_qn")

# write tables
table(de_none_q3$contrast, de_none_q3$de)

```

## DE Volcano plots

```{r de volcano plots}

de <- NULL
de <- de_qq_qn

plot_de_volcano <- function(res, nlabel=50) {
  # label the top n up/down genes
  label_data <- bind_rows(filter(res, de == "up") %>% slice_max(log2fc, n=nlabel),
                          filter(res, de == "dn") %>% slice_min(log2fc, n=nlabel))
  label_data <- distinct(label_data)

  p <- ggplot(res, aes(x=log2fc, y=-log10(padj + 1e-300), color=de, size=de)) +
    geom_point(alpha=0.7) +
    geom_text_repel(data=label_data, color="black", size=3, aes(label=gene), max.overlaps=Inf) +
    scale_color_manual(values=c("no"="grey", "dn"="blue", "up"="red")) +
    scale_size_manual(values=c("no"=1, "dn"=2, "up"=2)) +
    theme_minimal() +
    theme(legend.position="bottom") +
    theme(axis.line = element_line(color = "black")) +
    labs(x="log2fc", y="-log10(padj)")
  return(p)
}

volcano_dir <- file.path(plot_dir, "volcano")

# volcano plots
for (mycontrast in unique(de$contrast)) {
  print(mycontrast)
  x <- filter(de, contrast == mycontrast)
  p <- plot_de_volcano(x) + 
    labs(title="Volcano Plot", subtitle=mycontrast)
  f <- paste0("volcano_de_", mycontrast)
  ggsave(file.path(volcano_dir, paste0(f, ".pdf")), p, width=8, height=8)
  ggsave(file.path(volcano_dir, paste0(f, ".png")), p, width=8, height=8)
}

```

## DE Heatmaps

```{r de heatmaps, include=FALSE}

# heatmap functions
plot_pheatmap <- function(m, 
                          annot_col,
                          # annot_row, 
                          annot_colors,
                          filename, width, height,
                          xmin=-6, xmax=6) {
  pheatmap(m,
           scale = "none",
           color = colorRampPalette(c("navy", "white", "firebrick3"))(50),
           breaks = seq(xmin, xmax, length.out=51),
           annotation_col = annot_col,
           # annotation_row = annot_row,
           # annotation_colors = annot_colors,
           border_color = NA,
           cluster_rows = TRUE,
           cluster_cols = TRUE,
           clustering_distance_rows = "euclidean",
           clustering_distance_cols = "euclidean",
           clustering_method = "ward.D2",
#           cellwidth = 6,
#           cellheight = 6,
           fontsize = 6,
           fontsize_row = 6,
           fontsize_col = 6,
           filename = filename, 
           width = width,
           height = height)
}

# setup heatmap annotations
# column annotations: show clusters, study
# no row annotations
gene_annot <- fds_qq_qn$meta
# %>%
#   left_join(hpa_pdac, by=c("gene"="gene_symbol")) %>%
#   left_join(cptac_de_merged, by=c("gene"="gene_symbol")) %>%
#   mutate(de_cptac_rna = replace_na(de_cptac_rna, "na"),
#          de_cptac_prot = replace_na(de_cptac_prot, "na"))
# annot_row <- gene_annot %>%
#   select(gene) %>%
#   column_to_rownames("gene") %>%
#   as.data.frame()
annot_col <- column_to_rownames(fds_qq_qn$samples, "aoi") %>%
  select(study, cluster_name) %>%
  as.data.frame()

#
# IPMN clusters
#
# de gene params
padj_cutoff <- 0.01
log2fc_cutoff <- 3
ntop <- 10
contrasts <- c("clusterPanc", "clusterLn", "clusterKidney", "clusterColon", "clusterBrain")

g <- filter(de_qq_qn, contrast %in% contrasts,
            abs(log2fc) > log2fc_cutoff, 
            padj < padj_cutoff) %>%
  group_by(contrast, de) %>%
  slice_max(abs(log2fc), n=ntop) %>%
  ungroup() %>%
  distinct(gene) %>% 
  pull(gene)
ngenes <- length(g)
cluster_marker_genes <- g

# centered gene expression (subtract median)
s <- fds_qq_qn$samples
x <- log2_norm_cpm_qq_qn[g, s$aoi]
u <- apply(x, 1, mean)
x <- sweep(x, 1, u, FUN="-")
xmin <- -6
xmax <- 6

# heatmap
heatmap_dir <- file.path(plot_dir, "heatmap")

f <- file.path(plot_dir,  paste0("heatmap_de_cluster_markers_", ngenes, "_genes"))
h <- 6
w <- 6
p <- plot_pheatmap(x, 
                   annot_col[s$aoi,], 
                   annot_row[g,], 
                   annot_colors=scale_color_continuous(),
                   filename=paste0(f, ".png"), 
                   width=w, height=h,
                   xmin=xmin, xmax=xmax)
save_pheatmap_pdf(p, filename=paste0(f, ".pdf"), width=w, height=h)


```


## PCA feature plots for marker genes

```{r pca feature plot for each marker gene, include=FALSE}

s <- fds_qq_qn$samples
y <- log2_norm_cpm_qq_qn
feature_plot_genes <- c("MS4A1", "CPA1", "SNAP25", "IGHA1", "IGHG1")

for (gene in feature_plot_genes) {
  print(gene)
  x <- bind_cols(s, as_tibble(pca_res_qq_qn$x), gene=unlist(y[gene,]))
  p <- ggplot(x, aes(x=PC1, y=PC2, color=gene, shape=study)) +
    geom_point(alpha = 0.8, size=3) +
    scale_color_viridis_c() +
    theme_minimal() +
    theme(aspect.ratio=1) +
    labs(x="PC1", y="PC2", title=gene)
  f <- file.path(plot_dir, paste0("pca_featureplot_", gene))
  ggsave(paste0(f, ".pdf"), plot=p, width=5, height=5)
  ggsave(paste0(f, ".png"), plot=p, width=5, height=5)
}

```


## DE 2D scatter plots

```{r ipmn de 2d scatter plots}

plot_de_scatter2d <- function(tbl, xcontrast, ycontrast, nlabel=10, 
                              highlight_genes=NULL) {
  if (is.null(highlight_genes)) {
    highlight_genes <- c()
  }
  xde <- sym(paste0("de_", xcontrast))
  yde <- sym(paste0("de_", ycontrast))
  xlog2fc <- sym(paste0("log2fc_", xcontrast))
  ylog2fc <- sym(paste0("log2fc_", ycontrast))

  x <- tbl %>% 
    mutate(subtype = paste0("x", {{xde}}, "_", "y", {{yde}}),
           log2fc_abs_x = case_when({{xde}} == "no" ~ 0,
                                    TRUE ~ abs({{xlog2fc}})),
           log2fc_abs_y = case_when({{yde}} == "no" ~ 0,
                                    TRUE ~ abs({{ylog2fc}})),
           log2fc_abs_max = pmax(log2fc_abs_x, log2fc_abs_y))
  
  color_scale = c(xno_yno = "#888888",
                  xup_yup = "#FF0000",
                  xdn_ydn = "#0000aa",
                  xup_ydn = "#eeee00",
                  xdn_yup = "#ff00ff",
                  xup_yno = "#32cd32",
                  xdn_yno = "#670092",
                  xno_yup = "#ff6700",
                  xno_ydn = "#16ccc4")
  size_scale = c(xno_yno = 1, 
                 xup_yup = 2,
                 xdn_ydn = 2,
                 xup_ydn = 2,
                 xdn_yup = 2,
                 xup_yno = 2,
                 xdn_yno = 2,
                 xno_yup = 2,
                 xno_ydn = 2)
  
  annot_data <- filter(x, subtype != "xno_yno")
  none_data <- filter(x, subtype == "xno_yno")
  label_data <- annot_data %>% 
    group_by(subtype) %>%
    slice_max(log2fc_abs_max, n=nlabel) %>%
    ungroup()
  highlight_data <- filter(x, gene %in% highlight_genes)

  p <- ggplot(x, aes(x={{xlog2fc}}, 
                     y={{ylog2fc}},
                     color=subtype, 
                     size=subtype)) +
    geom_point(data=none_data, alpha=0.4) + 
    geom_point(data=annot_data, alpha=0.8) +
    geom_text_repel(data=label_data, color="black", size=3, aes(label=gene), max.overlaps=Inf) + 
    geom_point(data=highlight_data, pch=21, fill=NA, size=5, colour="#ffff00", stroke=1) + 
    geom_point(data=highlight_data, pch=21, fill=NA, size=4, colour="#222222", stroke=1, alpha=0.5) +
    geom_text_repel(data=highlight_data, color="black", size=3, aes(label=gene), max.overlaps=Inf) +     
    scale_color_manual(values = color_scale) +
    scale_size_manual(values = size_scale, guide = "none") + 
    theme_minimal() +
    xlab(rlang::as_string(xlog2fc)) +
    ylab(rlang::as_string(ylog2fc)) +
    labs(color = "DE Genes")
  return(p)
}


# merge de analysis by gene
de_merged <- de_qq_qn %>% select(gene, contrast, log2fc, padj, de) %>%
  pivot_wider(names_from = contrast, values_from = c(log2fc, padj, de))
nlabel <- 10
width <- 8
height <- 8

#
# IPMN Clusters
#
p <- plot_de_scatter2d(de_merged, 
                       xcontrast="clusterColon", 
                       ycontrast="clusterLn", 
                       nlabel=nlabel) +
  labs(title="IPMN Specimens", 
       subtitle="clusterColon (x-axis) clusterLn (y-axis)")
f <- file.path(plot_dir, "scatter2d_de_x_clusterColon_y_clusterLn")
ggsave(paste0(f, ".pdf"), plot=p, width=8, height=8)
ggsave(paste0(f, ".png"), plot=p, width=8, height=8)

```

## FIGURE 6: GSEA
``` {r gsea}
gsea_nperms <- 10000

read_msigdb_json_file <- function(filename) {
  x <- jsonlite::fromJSON(filename)
  y <- tibble(gs_name = names(x), gs = x) %>% 
    unnest_wider(gs) %>%
    unnest_longer(geneSymbols) %>%
    select(gs_cat = collection,
           gs_name = gs_name,
           gene_symbol = geneSymbols)
}

# read msigdb data: hallmark
f <- system.file("extdata", "h.all.v2024.1.Hs.json", package = "stgeomx")
msigdb_gene_sets <- read_msigdb_json_file(f)
# determine gene universe
msigdb_genes <- intersect(fds_qq_qn$meta$gene, msigdb_gene_sets$gene_symbol)
# filter genes not in our dataset
msigdb_gene_sets <- msigdb_gene_sets %>%
  dplyr::filter(gene_symbol %in% fds_qq_qn$meta$gene) %>%
  distinct(gs_cat, gs_name, gene_symbol, .keep_all = TRUE)
# gene set categories
gs_hallmark <- dplyr::filter(msigdb_gene_sets, gs_cat == "H")
gs_hallmark <- split(x = gs_hallmark$gene_symbol, f = gs_hallmark$gs_name)

# read msigdb data: pancreas_endocrine
f <- system.file("extdata", "GOBP_ENDOCRINE_PANCREAS_DEVELOPMENT.v2024.1.Hs.json", package = "stgeomx")
gobp_end_pancreas_gene_sets <- read_msigdb_json_file(f)
# determine gene universe
gobp_end_pancreas_genes <- intersect(fds_qq_qn$meta$gene, gobp_end_pancreas_gene_sets$gene_symbol)
# filter genes not in our dataset
gobp_end_pancreas_gene_sets <- gobp_end_pancreas_gene_sets %>%
  dplyr::filter(gene_symbol %in% fds_qq_qn$meta$gene) %>%
  distinct(gs_cat, gs_name, gene_symbol, .keep_all = TRUE)
# gene set categories
gs_end_panc <- dplyr::filter(gobp_end_pancreas_gene_sets, gs_cat == "C5:GO:BP")
gs_end_panc <- split(x = gs_end_panc$gene_symbol, f = gs_end_panc$gs_name)

# read msigdb data: pancreas secretion
f <- system.file("extdata", "GOBP_INSULIN_SECRETION.v2024.1.Hs.json", package = "stgeomx")
gobp_pancreas_secretion_gene_sets <- read_msigdb_json_file(f)
# determine gene universe
gobp_pancreas_secretion_genes <- intersect(fds_qq_qn$meta$gene, gobp_pancreas_secretion_gene_sets$gene_symbol)
# filter genes not in our dataset
gobp_pancreas_secretion_gene_sets <- gobp_pancreas_secretion_gene_sets %>%
  dplyr::filter(gene_symbol %in% fds_qq_qn$meta$gene) %>%
  distinct(gs_cat, gs_name, gene_symbol, .keep_all = TRUE)
# gene set categories
gs_panc_sec <- dplyr::filter(gobp_pancreas_secretion_gene_sets, gs_cat == "C5:GO:BP")
gs_panc_sec <- split(x = gs_panc_sec$gene_symbol, f = gs_panc_sec$gs_name)

# read msigdb data: brain development gene sets
f <- system.file("extdata", "brain_genesets.v2024.1.Hs.json", package = "stgeomx")
gobp_brain_gene_sets <- read_msigdb_json_file(f)
# determine gene universe
gobp_brain_genes <- intersect(fds_qq_qn$meta$gene, gobp_brain_gene_sets$gene_symbol)
# filter genes not in our dataset
gobp_brain_gene_sets <- gobp_brain_gene_sets %>%
  dplyr::filter(gene_symbol %in% fds_qq_qn$meta$gene) %>%
  distinct(gs_cat, gs_name, gene_symbol, .keep_all = TRUE)
# gene set categories
gs_brain <- dplyr::filter(gobp_brain_gene_sets, gs_cat == "C5:GO:BP")
gs_brain <- split(x = gs_brain$gene_symbol, f = gs_brain$gs_name)

# read msigdb data: ln development gene sets
f <- system.file("extdata", "ln_genesets.v2024.1.Hs.json", package = "stgeomx")
gobp_ln_gene_sets <- read_msigdb_json_file(f)
# determine gene universe
gobp_ln_genes <- intersect(fds_qq_qn$meta$gene, gobp_ln_gene_sets$gene_symbol)
# filter genes not in our dataset
gobp_ln_gene_sets <- gobp_ln_gene_sets %>%
  dplyr::filter(gene_symbol %in% fds_qq_qn$meta$gene) %>%
  distinct(gs_cat, gs_name, gene_symbol, .keep_all = TRUE)
# gene set categories
gs_ln <- dplyr::filter(gobp_ln_gene_sets, gs_cat == "C5:GO:BP")
gs_ln <- split(x = gs_ln$gene_symbol, f = gs_ln$gs_name)

# read msigdb data: kidney development gene sets
f <- system.file("extdata", "kidney_genesets.v2024.1.Hs.json", package = "stgeomx")
gobp_kidney_gene_sets <- read_msigdb_json_file(f)
# determine gene universe
gobp_kidney_genes <- intersect(fds_qq_qn$meta$gene, gobp_kidney_gene_sets$gene_symbol)
# filter genes not in our dataset
gobp_kidney_gene_sets <- gobp_kidney_gene_sets %>%
  dplyr::filter(gene_symbol %in% fds_qq_qn$meta$gene) %>%
  distinct(gs_cat, gs_name, gene_symbol, .keep_all = TRUE)
# gene set categories
gs_kidney <- dplyr::filter(gobp_kidney_gene_sets, gs_cat == "C5:GO:BP")
gs_kidney <- split(x = gs_kidney$gene_symbol, f = gs_kidney$gs_name)

# read msigdb data: colon development gene sets
f <- system.file("extdata", "colon_genesets.v2024.1.Hs.json", package = "stgeomx")
gobp_colon_gene_sets <- read_msigdb_json_file(f)
# determine gene universe
gobp_colon_genes <- intersect(fds_qq_qn$meta$gene, gobp_colon_gene_sets$gene_symbol)
# filter genes not in our dataset
gobp_colon_gene_sets <- gobp_colon_gene_sets %>%
  dplyr::filter(gene_symbol %in% fds_qq_qn$meta$gene) %>%
  distinct(gs_cat, gs_name, gene_symbol, .keep_all = TRUE)
# gene set categories
gs_colon <- dplyr::filter(gobp_colon_gene_sets, gs_cat == "C5:GO:BP")
gs_colon <- split(x = gs_colon$gene_symbol, f = gs_colon$gs_name)

# new gsea code
get_fgsea_ranks_de <- function(xde, mycontrast) {
  ranks <- xde %>%
    dplyr::filter(contrast == mycontrast) %>%
    select(gene, log2fc, padj) %>%
    mutate(rank = log2fc)
  ranks = sort(setNames(ranks$rank, ranks$gene), decreasing = TRUE)
  return(ranks)
}
 
 
fgsea_run_batch <- function(xde, xgs,
                            fgsea_eps=0,
                            fgsea_nperms=1000) {
  gsea <- NULL
  for (xcontrast in unique(xde$contrast)) {
    ranks <- get_fgsea_ranks_de(xde, xcontrast)
    res <- fgsea(pathways=xgs,
                 stats=ranks,
                 eps=fgsea_eps,
                 nPermSimple=fgsea_nperms)
    res <- mutate(res, contrast = xcontrast)
    gsea <- bind_rows(gsea, res)     
  }
  return(gsea)
}

stgeomx_gsea_pipeline <- function(gsea, de, method) {
  res <- fgsea_run_batch(de, gs_hallmark,
                       fgsea_eps=0, fgsea_nperms=gsea_nperms)
  res <- res %>% mutate(method = method)
  gsea <- bind_rows(gsea, res)
  
  res <- fgsea_run_batch(de, gs_ex_panc,
                         fgsea_eps=0, fgsea_nperms=gsea_nperms)
  res <- res %>% mutate(method = method)
  gsea <- bind_rows(gsea, res)
  
  
  res <- fgsea_run_batch(de, gs_end_panc,
                         fgsea_eps=0, fgsea_nperms=gsea_nperms)
  res <- res %>% mutate(method = method)
  gsea <- bind_rows(gsea, res)
  
  res <- fgsea_run_batch(de, gs_panc_sec,
                         fgsea_eps=0, fgsea_nperms=gsea_nperms)
  res <- res %>% mutate(method = method)
  gsea <- bind_rows(gsea, res)
  
  res <- fgsea_run_batch(de, gs_brain,
                         fgsea_eps=0, fgsea_nperms=gsea_nperms)
  res <- res %>% mutate(method = method)
  gsea <- bind_rows(gsea, res)
  
  res <- fgsea_run_batch(de, gs_ln,
                         fgsea_eps=0, fgsea_nperms=gsea_nperms)
  res <- res %>% mutate(method = method)
  gsea <- bind_rows(gsea, res)
  
  res <- fgsea_run_batch(de, gs_kidney,
                         fgsea_eps=0, fgsea_nperms=gsea_nperms)
  res <- res %>% mutate(method = method)
  gsea <- bind_rows(gsea, res)
  
  res <- fgsea_run_batch(de, gs_colon,
                         fgsea_eps=0, fgsea_nperms=gsea_nperms)
  res <- res %>% mutate(method = method)
  gsea <- bind_rows(gsea, res)
  
  # filter out irrelevant genes in hallmark
  gsea <- gsea %>%
  filter(!(pathway %in% c("HALLMARK_ESTROGEN_RESPONSE_LATE", "HALLMARK_ESTROGEN_RESPONSE_EARLY", "HALLMARK_GLYCOLYSIS", "HALLMARK_P53_PATHWAY", "HALLMARK_HEME_METABOLISM", "HALLMARK_ADIPOGENESIS", "HALLMARK_MYC_TARGETS_V2", "HALLMARK_MYC_TARGETS_V1", "HALLMARK_OXIDATIVE_PHOSPHORYLATION", "HALLMARK_HYPOXIA", "HALLMARK_UV_RESPONSE_DN", "HALLMARK_MYOGENESIS", "HALLMARK_APICAL_SURFACE", "HALLMARK_APOPTOSIS", "HALLMARK_IL6_JAK_STAT3_SIGNALING", "HALLMARK_IL2_STAT5_SIGNALING", "HALLMARK_MITOTIC_SPINDLE", "HALLMARK_APICAL_JUNCTION", "HALLMARK_KRAS_SIGNALING_UP", "HALLMARK_KRAS_SIGNALING_DN", "HALLMARK_UNFOLDED_PROTEIN_RESPONSE", "HALLMARK_DNA_REPAIR", "HALLMARK_FATTY_ACID_METABOLISM", "HALLMARK_HEDGEHOG_SIGNALING", "HALLMARK_ANGIOGENESIS", "HALLMARK_E2F_TARGETS", "HALLMARK_G2M_CHECKPOINT", "HALLMARK_TNFA_SIGNALING_VIA_NFKB", "HALLMARK_PROTEIN_SECRETION", "HALLMARK_CHOLESTEROL_HOMEOSTASIS")))
  
  return(gsea)
}

# Run full analysis
gsea_all <- NULL
gsea_none_q3 <- NULL
gsea_none_tmm <- NULL
gsea_none_rle <- NULL
gsea_none_qn <- NULL

gsea_bgsub_q3 <- NULL
gsea_bgsub_tmm <- NULL
gsea_bgsub_rle <- NULL
gsea_bgsub_qn <- NULL

gsea_qq_q3 <- NULL
gsea_qq_tmm <- NULL
gsea_qq_rle <- NULL
gsea_qq_qn <- NULL

# run stgeomx pipeline
gsea_none_q3 <- stgeomx_gsea_pipeline(gsea_none_q3, de_none_q3, "none_q3")
gsea_none_tmm <- stgeomx_gsea_pipeline(gsea_none_tmm, de_none_tmm, "none_tmm")
gsea_none_rle <- stgeomx_gsea_pipeline(gsea_none_rle, de_none_rle, "none_rle")
gsea_none_qn <- stgeomx_gsea_pipeline(gsea_none_qn, de_none_qn, "none_qn")

gsea_bgsub_q3 <- stgeomx_gsea_pipeline(gsea_bgsub_q3, de_bgsub_q3, "bgsub_q3")
gsea_bgsub_tmm <- stgeomx_gsea_pipeline(gsea_bgsub_tmm, de_bgsub_tmm, "bgsub_tmm")
gsea_bgsub_rle <- stgeomx_gsea_pipeline(gsea_bgsub_rle, de_bgsub_rle, "bgsub_rle")
gsea_bgsub_qn <- stgeomx_gsea_pipeline(gsea_bgsub_qn, de_bgsub_qn, "bgsub_qn")

gsea_qq_q3 <- stgeomx_gsea_pipeline(gsea_qq_q3, de_qq_q3, "qq_q3")
gsea_qq_tmm <- stgeomx_gsea_pipeline(gsea_qq_tmm, de_qq_tmm, "qq_tmm")
gsea_qq_rle <- stgeomx_gsea_pipeline(gsea_qq_rle, de_qq_rle, "qq_rle")
gsea_qq_qn <- stgeomx_gsea_pipeline(gsea_qq_qn, de_qq_qn, "qq_qn")

# # write results for clusters
# contrasts <- c("clusterPanc", "clusterBrain", "clusterLn", "clusterKidney", "clusterColon")
# manuscript_pathways <- c(names(gs_hallmark))
# res <- gsea_all %>% filter(contrast %in% contrasts, 
#                            pathway %in% manuscript_pathways)

# prefix <- "gsea_all"
# f <- file.path(results_dir, paste0(prefix, "_results.tsv"))
# data.table::fwrite(res, file=f, sep="\t", sep2=c("", " ", ""))
# f <- file.path(results_dir, paste0(prefix, "_results.xlsx"))
# write_xlsx(res, f)
```
### GSEA plots: helper functions
```{r helpers}
gsea_padj_cutoff <- 0.01
gsea_log2fc_cutoff <- 1

# contrasts
contrasts <- c("clusterPanc", "clusterBrain", "clusterLn", "clusterKidney", "clusterColon")

prep_gsea <- function(gsea) {
  # prepare gsea results
  gsea <- gsea %>%
    filter(contrast %in% contrasts) %>%
    mutate(padj = 2 * padj) %>%
    mutate(contrast = factor(contrast, levels=contrasts),
           neglog10padj = ifelse(padj < gsea_padj_cutoff, -log10(padj), NA),
           sig = padj < gsea_padj_cutoff) %>%
    group_by(contrast, pathway) %>%
    slice_min(padj, n=1, with_ties=FALSE) %>%
    ungroup()
  return(gsea)
}

prep_hallmark <- function(gsea) {
  # Hallmark
  res <- gsea %>% filter(pathway %in% names(gs_hallmark)) %>%
    mutate(pathway = gsub("HALLMARK_", "", pathway))
  # only keep significant results (at least one contrast)
  res <- res %>% group_by(pathway) %>%
    filter(any(padj < gsea_padj_cutoff)) %>%
    ungroup()
  return(res)
}

prep_all <- function(gsea) {
  # remove pathway prefixes
  res <- gsea  %>% mutate(pathway = str_replace(pathway, "^HALLMARK_|^GOBP_", ""))
  # only keep significant results (at least one contrast)
  res <- res %>% group_by(pathway) %>%
    filter(any(padj < gsea_padj_cutoff)) %>%
    ungroup()
  return(res)
}
```

### GSEA plots: by cluster type

```{r gsea plots}
cluster = "clusterColon"

plot_cluster <- function(res, pathway_levels, cluster) {
  res_panc <- res %>% filter(contrast == cluster)
  
  # Set the pathway levels to be the same as the reference (res_qq_qn)
  res_panc$pathway <- factor(res_panc$pathway, levels=pathway_levels)
  
  # Filter out any NA pathways
  res_panc <- res_panc %>% filter(!is.na(pathway))
  
  p <- ggplot(res_panc, aes(x=NES, y=pathway, fill=neglog10padj)) +
    geom_col() +
    scale_fill_viridis_c() +
    theme_bw() +
    theme(axis.text.y = element_text(color="black", size=7)) +
    theme(strip.text = element_text(size=8)) +
    facet_wrap(~ contrast, nrow=1) +
    labs(x="Normalized Enrichment Score (NES)", y="Pathway")
  
  return (p)
}

# Prepare and process qq_qn data
gsea_qq_qn <- prep_gsea(gsea_qq_qn)
res_qq_qn <- prep_all(gsea_qq_qn)

# Extract pathway levels from res_qq_qn to use for all plots
pathway_levels <- res_qq_qn %>% filter(contrast == cluster) %>% arrange(NES) %>% pull(pathway)

# Generate qq_qn plot using consistent pathway levels
p_qq_qn <- plot_cluster(res_qq_qn, pathway_levels, cluster)

# Prepare and process none_q3 data
gsea_none_q3 <- prep_gsea(gsea_none_q3)
res_none_q3 <- prep_all(gsea_none_q3)
p_none_q3 <- plot_cluster(res_none_q3, pathway_levels, cluster)

gsea_none_tmm <- prep_gsea(gsea_none_tmm)
res_none_tmm <- prep_all(gsea_none_tmm)
p_none_tmm <- plot_cluster(res_none_tmm, pathway_levels, cluster)

gsea_none_rle <- prep_gsea(gsea_none_rle)
res_none_rle <- prep_all(gsea_none_rle)
p_none_rle <- plot_cluster(res_none_rle, pathway_levels, cluster)

gsea_none_qn <- prep_gsea(gsea_none_qn)
res_none_qn <- prep_all(gsea_none_qn)
p_none_qn <- plot_cluster(res_none_qn, pathway_levels, cluster)

gsea_bgsub_q3 <- prep_gsea(gsea_bgsub_q3)
res_bgsub_q3 <- prep_all(gsea_bgsub_q3)
p_bgsub_q3 <- plot_cluster(res_bgsub_q3, pathway_levels, cluster)

gsea_bgsub_tmm <- prep_gsea(gsea_bgsub_tmm)
res_bgsub_tmm <- prep_all(gsea_bgsub_tmm)
p_bgsub_tmm <- plot_cluster(res_bgsub_tmm, pathway_levels, cluster)

gsea_bgsub_rle <- prep_gsea(gsea_bgsub_rle)
res_bgsub_rle <- prep_all(gsea_bgsub_rle)
p_bgsub_rle <- plot_cluster(res_bgsub_rle, pathway_levels, cluster)

gsea_bgsub_qn <- prep_gsea(gsea_bgsub_qn)
res_bgsub_qn <- prep_all(gsea_bgsub_qn)
p_bgsub_qn <- plot_cluster(res_bgsub_qn, pathway_levels, cluster)


# Prepare and process qq
gsea_qq_q3 <- prep_gsea(gsea_qq_q3)
res_qq_q3 <- prep_all(gsea_qq_q3)
p_qq_q3 <- plot_cluster(res_qq_q3, pathway_levels, cluster)

gsea_qq_tmm <- prep_gsea(gsea_qq_tmm)
res_qq_tmm <- prep_all(gsea_qq_tmm)
p_qq_tmm <- plot_cluster(res_qq_tmm, pathway_levels, cluster)

gsea_qq_rle <- prep_gsea(gsea_qq_rle)
res_qq_rle <- prep_all(gsea_qq_rle)
p_qq_rle <- plot_cluster(res_qq_rle, pathway_levels, cluster)

gsea_qq_qn <- prep_gsea(gsea_qq_qn)
res_qq_qn <- prep_all(gsea_qq_qn)
p_qq_qn <- plot_cluster(res_qq_qn, pathway_levels, cluster)
```
 
### GSEA combined plot: cluster type
``` {r GSEA combined plot}
gsea_dir <- file.path(plot_dir, "gsea")

# Grid column and row titles
col_titles <- list(
  wrap_elements(grid::textGrob('', gp = grid::gpar(fontsize = 18, fontface = "bold"))),
  wrap_elements(grid::textGrob('Q3', gp = grid::gpar(fontsize = 18, fontface = "bold"))),
  wrap_elements(grid::textGrob('TMM', gp = grid::gpar(fontsize = 18, fontface = "bold"))),
  wrap_elements(grid::textGrob('RLE', gp = grid::gpar(fontsize = 18, fontface = "bold"))),
  wrap_elements(grid::textGrob('QN', gp = grid::gpar(fontsize = 18, fontface = "bold")))
)

row_titles <- list(
  wrap_elements(grid::textGrob('NONE', gp = grid::gpar(fontsize = 18, fontface = "bold"))),
  wrap_elements(grid::textGrob('BGSub', gp = grid::gpar(fontsize = 18, fontface = "bold"))),
  wrap_elements(grid::textGrob('QQ', gp = grid::gpar(fontsize = 18, fontface = "bold")))
)

# Create grid
p <- wrap_plots(
  col_titles[[1]], col_titles[[2]], col_titles[[3]], col_titles[[4]], col_titles[[5]],
  row_titles[[1]], p_none_q3, p_none_tmm, p_none_rle, p_none_qn,
  row_titles[[2]], p_bgsub_q3, p_bgsub_q3, p_bgsub_q3, p_bgsub_q3,
  row_titles[[3]], p_qq_q3, p_qq_q3, p_qq_q3, p_qq_qn,
  ncol = 5,
  heights = c(0.1, 1, 1, 1),
  widths = c(0.3, 1, 1, 1, 1)
) +
plot_layout(
  guides = 'collect'
) &
theme(plot.margin = unit(c(0.5, 1, 1, 1), "cm"))

# Add titles
p <- p + 
  plot_annotation(
    title = "Effect of Background Correction and Normalization Methods on PCA"
  ) & 
  theme(
    plot.title = element_text(size = 22, face = "bold", hjust = 0.5)
  )

f <- file.path(gsea_dir, "gsea_colon")
ggsave(paste0(f, ".png"), p, width=35, height=20)
ggsave(paste0(f, ".pdf"), p, width=35, height=20)
```


### GSEA plots: by bg + normalization
``` {r gsea bg_norm}
gsea_dir <- file.path(plot_dir, "gsea")

contrasts <- c("clusterPanc", "clusterBrain", "clusterLn", "clusterKidney", "clusterColon")

plot_clusters <- function(res) {
  res_panc <- res %>% filter(contrast == "clusterPanc")
  pathway_levels <- res_panc %>% filter(contrast == "clusterPanc") %>% arrange(NES) %>% pull(pathway)
  res_panc$pathway <- factor(res_panc$pathway, levels=pathway_levels)
  
  res_brain <- res %>% filter(contrast == "clusterBrain")
  pathway_levels <- res_brain %>% filter(contrast == "clusterBrain") %>% arrange(NES) %>% pull(pathway)
  res_brain$pathway <- factor(res_brain$pathway, levels=pathway_levels)
  
  res_ln <- res %>% filter(contrast == "clusterLn")
  pathway_levels <- res_ln %>% filter(contrast == "clusterLn") %>% arrange(NES) %>% pull(pathway)
  res_ln$pathway <- factor(res_ln$pathway, levels=pathway_levels)
  
  res_kidney <- res %>% filter(contrast == "clusterKidney")
  pathway_levels <- res_kidney %>% filter(contrast == "clusterKidney") %>% arrange(NES) %>% pull(pathway)
  res_kidney$pathway <- factor(res_kidney$pathway, levels=pathway_levels)
  
  res_colon <- res %>% filter(contrast == "clusterColon")
  pathway_levels <- res_colon %>% filter(contrast == "clusterColon") %>% arrange(NES) %>% pull(pathway)
  res_colon$pathway <- factor(res_colon$pathway, levels=pathway_levels)
  
  p_panc <- ggplot(res_panc, aes(x=NES, y=pathway, fill=neglog10padj)) +
    geom_col() +
    scale_fill_viridis_c() +
    theme_bw() +
    theme(axis.text.y = element_text(color="black", size=7)) +
    theme(strip.text = element_text(size=8)) +
    facet_wrap(~ contrast, nrow=1) +
    labs(x="Normalized Enrichment Score (NES)", y="Pathway")
  p_brain <- ggplot(res_brain, aes(x=NES, y=pathway, fill=neglog10padj)) +
    geom_col() +
    scale_fill_viridis_c() +
    theme_bw() +
    theme(axis.text.y = element_text(color="black", size=7)) +
    theme(strip.text = element_text(size=8)) +
    facet_wrap(~ contrast, nrow=1) +
    labs(x="Normalized Enrichment Score (NES)", y="Pathway")
  p_ln <- ggplot(res_ln, aes(x=NES, y=pathway, fill=neglog10padj)) +
    geom_col() +
    scale_fill_viridis_c() +
    theme_bw() +
    theme(axis.text.y = element_text(color="black", size=7)) +
    theme(strip.text = element_text(size=8)) +
    facet_wrap(~ contrast, nrow=1) +
    labs(x="Normalized Enrichment Score (NES)", y="Pathway")
  p_kidney <- ggplot(res_kidney, aes(x=NES, y=pathway, fill=neglog10padj)) +
    geom_col() +
    scale_fill_viridis_c() +
    theme_bw() +
    theme(axis.text.y = element_text(color="black", size=7)) +
    theme(strip.text = element_text(size=8)) +
    facet_wrap(~ contrast, nrow=1) +
    labs(x="Normalized Enrichment Score (NES)", y="Pathway")
  p_colon <- ggplot(res_colon, aes(x=NES, y=pathway, fill=neglog10padj)) +
    geom_col() +
    scale_fill_viridis_c() +
    theme_bw() +
    theme(axis.text.y = element_text(color="black", size=7)) +
    theme(strip.text = element_text(size=8)) +
    facet_wrap(~ contrast, nrow=1) +
    labs(x="Normalized Enrichment Score (NES)", y="Pathway")
  
  return (p_panc + p_brain + p_ln + p_kidney + p_colon)
}

gsea_none_q3 <- prep_gsea(gsea_none_q3)
res_none_q3 <- prep_all(gsea_none_q3)
p <- plot_clusters(res_none_q3)
f <- file.path(gsea_dir, "gsea_none_q3")
ggsave(paste0(f, ".png"), p, width=20, height=15)
ggsave(paste0(f, ".pdf"), p, width=20, height=15)

gsea_none_tmm <- prep_gsea(gsea_none_tmm)
res_none_tmm <- prep_all(gsea_none_tmm)
p <- plot_clusters(res_none_tmm)
f <- file.path(gsea_dir, "gsea_none_tmm")
ggsave(paste0(f, ".png"), p, width=20, height=15)
ggsave(paste0(f, ".pdf"), p, width=20, height=15)

gsea_none_rle <- prep_gsea(gsea_none_rle)
res_none_rle <- prep_all(gsea_none_rle)
p <- plot_clusters(res_none_rle)
f <- file.path(gsea_dir, "gsea_none_rle")
ggsave(paste0(f, ".png"), p, width=20, height=15)
ggsave(paste0(f, ".pdf"), p, width=20, height=15)

gsea_none_qn <- prep_gsea(gsea_none_qn)
res_none_qn <- prep_all(gsea_none_qn)
p <- plot_clusters(res_none_qn)
f <- file.path(gsea_dir, "gsea_none_qn")
ggsave(paste0(f, ".png"), p, width=20, height=15)
ggsave(paste0(f, ".pdf"), p, width=20, height=15)

gsea_bgsub_q3 <- prep_gsea(gsea_bgsub_q3)
res_bgsub_q3 <- prep_all(gsea_bgsub_q3)
p <- plot_clusters(res_bgsub_q3)
f <- file.path(gsea_dir, "gsea_bgsub_q3")
ggsave(paste0(f, ".png"), p, width=20, height=15)
ggsave(paste0(f, ".pdf"), p, width=20, height=15)

gsea_bgsub_tmm <- prep_gsea(gsea_bgsub_tmm)
res_bgsub_tmm <- prep_all(gsea_bgsub_tmm)
p <- plot_clusters(res_bgsub_tmm)
f <- file.path(gsea_dir, "gsea_bgsub_tmm")
ggsave(paste0(f, ".png"), p, width=20, height=15)
ggsave(paste0(f, ".pdf"), p, width=20, height=15)

gsea_bgsub_rle <- prep_gsea(gsea_bgsub_rle)
res_bgsub_rle <- prep_all(gsea_bgsub_rle)
p <- plot_clusters(res_bgsub_rle)
f <- file.path(gsea_dir, "gsea_bgsub_rle")
ggsave(paste0(f, ".png"), p, width=20, height=15)
ggsave(paste0(f, ".pdf"), p, width=20, height=15)

gsea_bgsub_qn <- prep_gsea(gsea_bgsub_qn)
res_bgsub_qn <- prep_all(gsea_bgsub_qn)
p <- plot_clusters(res_bgsub_qn)
f <- file.path(gsea_dir, "gsea_bgsub_qn")
ggsave(paste0(f, ".png"), p, width=20, height=15)
ggsave(paste0(f, ".pdf"), p, width=20, height=15)

gsea_qq_q3 <- prep_gsea(gsea_qq_q3)
res_qq_q3 <- prep_all(gsea_qq_q3)
p <- plot_clusters(res_qq_q3)
f <- file.path(gsea_dir, "gsea_qq_q3")
ggsave(paste0(f, ".png"), p, width=20, height=15)
ggsave(paste0(f, ".pdf"), p, width=20, height=15)

gsea_qq_tmm <- prep_gsea(gsea_qq_tmm)
res_qq_tmm <- prep_all(gsea_qq_tmm)
p <- plot_clusters(res_qq_tmm)
f <- file.path(gsea_dir, "gsea_qq_tmm")
ggsave(paste0(f, ".png"), p, width=20, height=15)
ggsave(paste0(f, ".pdf"), p, width=20, height=15)

gsea_qq_rle <- prep_gsea(gsea_qq_rle)
res_qq_rle <- prep_all(gsea_qq_rle)
p <- plot_clusters(res_qq_rle)
f <- file.path(gsea_dir, "gsea_qq_rle")
ggsave(paste0(f, ".png"), p, width=20, height=15)
ggsave(paste0(f, ".pdf"), p, width=20, height=15)

gsea_qq_qn <- prep_gsea(gsea_qq_qn)
res_qq_qn <- prep_all(gsea_qq_qn)
p <- plot_clusters(res_qq_qn)
f <- file.path(gsea_dir, "gsea_qq_qn")
ggsave(paste0(f, ".png"), p, width=20, height=15)
ggsave(paste0(f, ".pdf"), p, width=20, height=15)
```
